---
title: "Notebook 03 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)

theme_set(theme_minimal())

food <- read_csv(file.path("data", "food.csv"))
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

# Aesthetics and Scales

## Additional Aesthetics

In the introduction to the grammar of graphics, we saw that visualizations can
be built out of graphics layers. Each layer, in turn, is described by a data
set, a geometry, and a series of aes (aesthetic) mappings between variables and
features of the layer. The point geometry required `x` and `y` aesthetics; the
text and text repel layers also required a `label` aesthetic.

In addition to the required aesthetics, each geometry type also has a number of
optional aesthetics that we can use to add additional information to the plot.
For example, most geoms have a `color` aesthetic. The syntax for describing
this is exactly the same as with the required aesthetics; we place the name of
the aesthetic followed by the name of the associated variable name. Let's see
what happens when add a color aesthetic this to our scatterplot by relating
the variable `food_group` to the aes `color`:

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = total_fat, color = food_group))
```

Notice that R has done a lot of work for us. It determined all of the food
groups in the data set, assigned each to a color, built a legend, and modified
the points on the plot so that the colors align with the food groups. Can you
now tell what types of food have a large number of calories and fat? Which
kinds of food have the lowest calories and fat? What is the biggest difference
between fruits and vegetables from the plot?

Similarly, we can modify the size of the points according to a variable in the
data set by setting the `size` aesthetic. Here, we will make points larger or
smaller based on the saturated fat in each food item:

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = total_fat, size = sat_fat))
```

Both size and color can also be specified for the text, text repel, and line
geometries. There are a few other aesthetics that will be useful, and that we
will introduce as needed.

## Fixed Aesthetics

In the previous section we changed the default aes value for the color and size
of points by associating these to a variable in the data set. In the plots from
the last notebook, where we did not specify color or size, R choose a default
value for these: the color "black" and the size 1. What if we want to change
the defaults to a different fixed value? This can be done relatively easily,
but take care with the details because this is a common source of confusing
errors for users new to the grammar of graphics.

To change an aes to a fixed value, we specify the changed value inside the
`geom_` function, but **after** the `aes(` function. Here, for example, is how
we change the size of all the points to 4 (four times larger than the default):

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = total_fat), size = 4)
```

We can do the same with colors, but notice that we need to put the color name
inside of quotes:

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = total_fat), color = "pink")
```

For a list of available color names, see
[this file](http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf).
You can interchange the fixed and variable aes commands, and the relative order
should not effect the output. Just be sure the put fixed terms after closing
the aes command.

## Scales

In the examples above, R handles the specifics of how to pick the specific
colors and sizes. R has also figured how how large to make the axes, where
to add tick marks, and where to draw grid lines. Letting R deal with these
details is convenient because it frees us up to focus on the data itself.
Sometimes, such as when preparing to produce plots for external distribution,
it is useful to manually adjust these details. That is exactly what scales are
for!

Each aesthetic within the grammar of graphics is associated with a scale.
Scales detail how a plot should relate aesthetics to the concrete, perceivable  
features in a plot. For example, a scale for the `x` aesthetic will describe
the smallest and largest values on the x-axis. A color scale would describe
what color corresponds to each category in a data set.

In order to change or modify the default scales, we add an additional item
to the ggplot code. The order of the scales relative to the geoms does not
effect the output; by convention, scales are usually grouped after the
geometries.

For example, a popular alternative to the default color palette shown above is
the function `scale_color_viridis_d()`. It constructs a set of colors that is:
(1) color-blind friendly, (2) looks nice when printed in black and white, and
(3) still displays fine on bad projectors. To use it, add the function
`scale_color_viridis_d` on as an extra row to the plot:

```{r, message=FALSE}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = sat_fat, color = food_group)) +
    scale_color_viridis_d()
```

There is also `scale_color_viridis_c` that produces a similar set of colors
when you want to color point according to a continuous variable.

```{r, message=FALSE}
food %>%
  ggplot() +
    geom_point(aes(x = total_fat, y = sat_fat, color = calories)) +
    scale_color_viridis_c()
```

Similarly, there are several scales to control the x and y axes. For example,
we can add `scale_x_log10()` and `scale_y_log10()` to the plot to produce a
plot on a logarithmic scale, which can be very useful when working with
heavily skewed data sets:

```{r, message=FALSE}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = sat_fat)) +
    scale_x_log10() +
    scale_y_log10()
```

The functions `scale_x_continuous()` and `scale_y_continuous()` reproduce the
default axes, but are useful for setting other details such as where to draw
horizontal and vertical guidelines.

## Extra color scales

There are several special scale types that can be useful for working with
colors. In some cases we may already have a column in our data set that
explicitly describes the color of an observations. This is, in fact, the case
with the food data set. In this case, we may want to use these colors directly.
To do that, use the scale `scale_color_identity`. Here is an example with each
food colored according to its assigned color:

```{r, message=FALSE}
food %>%
  ggplot() +
    geom_text_repel(
      aes(x = calories, y = sugar, color = color, label = item)
    ) +
    scale_color_identity()
```

Notice that by default no legend is created for the scale.

Another type of scale that can be useful for colors is `scale_color_manual`.
Here, it is possible to describe exactly which color should be used for each
category. Here is the syntax, with manually defined colors for each food group:

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = sugar, color = food_group)) +
    scale_color_manual(values = c(
      dairy = "lightblue",
      fish = "navy",
      fruit = "peachpuff1",
      grains = "wheat",
      meat = "indianred1",
      vegetable = "green"
    ))
```

Using manual colors is generally advisable in the case where there are
well-known colors associated with the groups in the data set. For example,
when plotting data about political parties it may make be helpful to use the
colors traditionally associated with each party.

## Inheritance of aesthetics

As a final optional point, note that there is a convention for simplifying the
plotting command. Often, each layer will use the same x and y variables. It is
possible to specify these just once in the `ggplot` function, and they will be
used by default in all other layers. Also, you can drop the `x =` and `y =` if
you put these options first. Here is an example of layering together
the `geom_point` and `geom_text_repel` with this inheritance structure:

```{r}
food %>%
  ggplot(aes(calories, total_fat)) +
    geom_point() +
    geom_text_repel(aes(label = item))
```

These changes are optional however, and you can feel free to write them as we
did earlier if you prefer. It is important to be able to recognize them,
though, if you are searching through documentation or help pages.

# Practice

For practice, we will work again with the largest cities datasets:

```{r, message=FALSE}
cities <- read_csv(file.path("data", "largest_cities.csv"))
```

We will also work with a second dataset of U.S. city populations over time.
The code below, which we will learn about next time, will select three
cities of data (Boston, Chicago, and LA) for us to work with:

```{r, message=FALSE}
us_three <- read_csv(file.path("data", "us_city_population.csv")) %>%
  filter(city %in% c("Boston, MA", "Chicago, IL", "Los Angeles, CA"))
```

You may want to take a moment to look at this new dataset to understand what
variables are available.

## Color and Size

In this section we will use the `cities` dataset. In the code below, produce a
scatter plot with longitude on the x-axis and latitude on the y-axis. Make the
points scale with the size of the city's population.

```{r}
cities %>%
  ggplot() +
    geom_point(aes(x = lon, y = lat, size = population))
```

What region(s) seem to have the highest concentration of large cities?
**Answer**: Most of the largest cities are in Indian and China.

Below, repeat the same plot but now scale the color (not the size) of the
points based on the population.

```{r}
cities %>%
  ggplot() +
    geom_point(aes(x = lon, y = lat, color = population))
```

Which of these two plots (color or size) do you find easier to read? Why?
**Answer**: With the default scale, I find the size plot much easier to
read because the default continuous scale is really quite bad.

In the code below, plot the cities data with the city longitude on the x-axis
and latitude on the y-axis. Color the points based on the continent the city
is in and make the size proportional to the population.

```{r}
cities %>%
  ggplot() +
    geom_point(aes(x = lon, y = lat, color = continent, size = population))
```

What "holes" in the map exist? That is, what are some large regions have no
cities in the dataset that you tend to expect to find on a map?
**Answer**: There are no large cities in the region that I belive Russia
occupies, nor in Northern Africa.

## Group Aesthetic

In this section, we will use the `us_three` dataset. In the code below we are
going to produce a line plot with year on the x-axis and population on the
y-axis. To do this, use the `geom_line` geometry.

```{r}
us_three %>%
  ggplot() +
    geom_line(aes(x = year, y = population))
```

You should see a strange plot that seems to have a jagged, comb-like pattern to
it. Below, repeat this plot but add a points layer that is colored by the city
name (do not add color yet to the lines)

```{r}
us_three %>%
  ggplot() +
    geom_line(aes(x = year, y = population)) +
    geom_point(aes(x = year, y = population, color = city))
```

Can you see what's happening now? The line plot is drawing a single line,
from the smallest year to the largest year, without ever lifting up its pen.
It is connecting across time and space, producing a plot that is nearly
meaningless (or at least, very hard to interpret).

In the code below, produce a line plot with year on the x-axis and population
on the y-axis. Color the lines by city. Do not include a points layer.

```{r}
us_three %>%
  ggplot() +
    geom_line(aes(x = year, y = population, color = city))
```

This plot should now look a lot nicer as it shows a different line for each
city. To understand what is going on, produce the same plot in the code below,
but replace the `color` aes with a new aes called `group`. In other words, map
the `group` aesthetic to the city aesthetic.

```{r}
us_three %>%
  ggplot() +
    geom_line(aes(x = year, y = population, group = city))
```

You should now see three lines that are all in black. It is this exact plot
that was being colored in the previous example.

What's going on here? When we add aesthetics such as color and size to a line
geometry, another aesthetic called `group` is magically set at the same time.
It is possible to set this aesthetic explicitly if we want to group the data
without producing a different size or color.

## Fixed Aesthetics

Now, return again to the `cities` dataset. Produce a scatter plot with
longitude and latitude, but color all of the points the color "pink".

```{r}
cities %>%
  ggplot() +
    geom_point(aes(x = lon, y = lat), color = "pink")
```

Starting from the plot in the previous question, modify the size of the points
to change with the population of each city.

```{r}
cities %>%
  ggplot() +
    geom_point(aes(x = lon, y = lat, size = population), color = "pink")
```

Another aesthetic value that was not covered in the notes above is `alpha`.
It is given by a number between 0 and 1 and sets how opaque the points are
shown. Most typically it is used as a fixed aesthetic to make a plot with a
large number of points more readable. Starting with the plot in the previous
question, add a fixed aesthetic to change the alpha value of the points to 0.4.

```{r}
cities %>%
  ggplot() +
    geom_point(
      aes(x = lon, y = lat, size = population),
      color = "pink",
      alpha = 0.4
    )
```

Notice how the areas where multiple cities overlap are darker than the
surrounding regions.

## Scales

With the `cities` dataset, produce a scatter plot with city area (`city_area`)
on the x-axis and city population (`city_pop`) on the y-axis.

```{r, warning=FALSE}
cities %>%
  ggplot() +
    geom_point(aes(x = city_area, y = city_pop))
```

You should find that it is difficult to interpret this plot because most of the
cities are squished together on the left-side of the plot. One way to fix this
is to use a logarithmic scale. In the code block below, provide a modification
of this plot with logarithmic scales for the x- and y-axes.

```{r, warning = FALSE}
cities %>%
  ggplot() +
    geom_point(aes(x = city_area, y = city_pop)) +
    scale_x_log10() +
    scale_y_log10()
```

You should find this plot easier to work with. If you are not familiar with
the scientific notation used on the x-axis labels here, make sure to
[learn more about these](https://en.wikipedia.org/wiki/Scientific_notation)
as they will appear from time-to-time throughout the notes in this course.

## Color Scales

The Köppen climate classification is a way of describing the climate of a region
or city. Information about these scores are included in the `cities` dataset and
are something that we will explore in length in future notebooks. Here we will
work with the four level categorization stored in the variable `koppen_main`
(there are actually five levels, but the coldest level is not present in this
dataset).

In the code block below, produce a scatter plot of the `cities` dataset with
longitude on the x-axis  and latitude on the y-axis. Color the points based on
the `koppen_main` variable.

```{r}
cities %>%
  ggplot() +
    geom_point(aes(x = lon, y = lat, color = koppen_main))
```

In the code below, modify this plot to use the color-blind friendly Viridis
color scale:

```{r}
cities %>%
  ggplot() +
    geom_point(aes(x = lon, y = lat, color = koppen_main)) +
    scale_color_viridis_d()
```

There is a well-known and standardized color scale for the Köppen codes. We will
use the full palette in a future notebook; here we will just set values for the
four main categories. In the code below repeat the scatterplot, but add a manual
color scale where the "Tropical" climates are blue, the "Dry" climates are red,
"Temperate" is green and "Continental" is purple.

```{r}
cities %>%
  ggplot() +
    geom_point(aes(x = lon, y = lat, color = koppen_main)) +
    scale_color_manual(values = c(
      Tropical = "blue",
      Dry = "red",
      Temperate = "green",
      Continental = "purple"
    ))
```
