---
title: "Notebook 02 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old datasets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)

theme_set(theme_minimal())

food <- read_csv(file.path("data", "food.csv"))
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

## Pipes

We are going to make extensive use of an operator called a *pipe* throughout
the course. Let's take a moment to understand what this operator does. We will
do this using an admittedly silly example, but the idea will make more sense as
we see some examples.

Consider the task of taking the cosine of the base-2 logarithm of the
square-root of the sine of the number 20. We can compute this in R by nesting
each of these functions inside of one another:

```{r}
cos(log(sqrt(sin(20)), base = 2))
```

Alternatively we could split each of these operations into their own line using
a temporary variable and the assignment operator:

```{r}
result <- sin(20)
result <- sqrt(result)
result <- log(result, base = 2)
result <- cos(result)
result
```

This is a bit easier to read and allows us to apply any number of functions with
complex optional parameters without running out of space. The downside is that
we need to create, name, and keep track of temporary variable names. The pipe
operator, written `%>%`, allows us to get the benefits of both techniques by
passing the output of one function to the first argument of the next function.
It works like this:

```{r}
20 %>%
  sin() %>%
  sqrt() %>%
  log(base = 2) %>%
  cos()
```

This may seem like overkill here, but it will be very useful in the code to
come.

## Grammar of Graphics

### Elements

Data visualization is an incredibly important skill and a task that R is
particularly well-designed for. We are going to learn and use the **ggplot2**
package for building beautiful and informative graphics. The package makes it
easy to build fairly complex graphics in a way that is guided by a general
theory of data visualization. The only downside is that, because it is built
around a theoretical model rather than many one-off solutions for different
tasks, it has a steep initial learning curve. These notes will, hopefully,
make this as painless as possible.

The core idea of the grammar of graphics is that visualizations are composed
of independent *layers*. To describe a specific layer, we need to specify
several elements:

- *data*: the dataset from which data will be taken to construct the plot
- *geom*: a description of what kinds of objects to plot (i.e., points, labels,
or boxes)
- *aes*: a mapping from elements of the plot to columns in our data set (i.e.,
the position on the x-axis or the color of our points); it stands for aesthetics

You can describe virtually any type of visualization by putting together these
elements.

To show how to use the grammar of graphics, we will start by using the food
dataset introduced in the previous notes, with each row describing a particular
item of food along with various nutritional information. The first plot we will
make is a scatter plot that investigates the relationship between calories and
the total fat (in grams) that are in a 100g portion of each food item. In the
language of the grammar of graphics we can describe this with the following
elements:

- *data*: our data set is called `food`
- *geom*: we will build a plot with a points geometry; each row of data is
represented by a point
- *aes*: the x-axis will be associated with `calories` and the y-axis with
`total_fat`

### Scatter plot example

The easiest way to understand how we specify these elements within **ggplot**
is by seeing an example. Here is the code to specify the data, geom, and aes:

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = total_fat))
```

In the first line we specify the data set (`food`), which is then piped (`%>%`)
into the function `ggplot`, which instructs R to start a new plot. Next, we add
(`+`) a layer to the plot. This layer uses a points geom (`geom_point`) and
describes two aes values, `x = calories` and `y = total_fat`.

In order to make a similar plot with different variables, or a different
dataset, you can copy this code and change the associated variable names
(`food`, `calories`, and `total_fat`). In the code below create another
scatterplot from the food dataset, choosing any two variables for the two
axes:

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = vitamin_a, y = iron))
```

In the next few classes we will see how to modify and build on this basic
structure to create more complex graphics.

### Text Geometries

Let's go through several other choices of geometries that we could have in the
plot. There are many of these, but in general you can create most plots with
only a small number of geometry types. To start, we will use the geometry
`geom_text`, which puts a small label in place of the points in our previous
example.

The text geometry needs an additional aes called `label` to describe what
variable in the dataset should be used as the label. Here, we use the variable
called `item` to label each point with the name of the specific food item in
question (the column is called `item`):

```{r}
food %>%
  ggplot() +
    geom_text(aes(x = calories, y = total_fat, label = item))
```

Can you now identify what food has the highest amount of fat? Or the highest
calorie count? Hopefully!

You likely cannot, however, figure out what foods have the lowest amount of fat
because the labels become too clumped together. In order to try to address this
issue, we can use a slightly different geometry called `geom_text_repel`. It
also places labels on the plot, but has logic that avoids intersecting labels.
Instead, labels are moved away from the data points and connected (when needed)
by a line segment:

```{r}
food %>%
  ggplot() +
    geom_text_repel(aes(x = calories, y = total_fat, label = item))
```

This is still a bit busy in the lower left-hand corner, but should be slightly
easier to read in the middle of the plot.

We can make the plot a bit more readable by adding two *layers*, one with
the text and another with the points. To do this, just add the two geometries
together like this:

```{r}
food %>%
  ggplot() +
    geom_text_repel(aes(x = calories, y = total_fat, label = item)) +
    geom_point(aes(x = calories, y = total_fat))
```

Next class we will see how to further improve this plot.

### Formatting code

The first notebook stress the importance of following a few style guidelines
about your code. Here are three additional formatting rules that apply
specifically to building graphics in R:

- indent every line that follows a pipe with two extra spaces
- indent every line below `ggplot` by an additional two spaces (four total)

As with our original set of style guidelines, you will make your life a lot
easier if you get used to these rules right from the start. Note that hitting
TAB should give you two spaces in the RStudio Cloud editor.

## Practice

### Cities Data

To further practice the Grammar of Graphics, we will work with the largest
cities data set.

```{r, message=FALSE}
cities <- read_csv(file.path("data", "largest_cities.csv"))
cities
```

Population numbers are given in millions and areas are given in square
kilometers. The three different kinds of populations refer to different
ways of describing what is included in a city. The variable `population` is a
measurement from the UN (2018) that tries to consistently define the region of
a city. More information is available in the U.N.'s World Urbanization
Prospectus, which is available
[here](https://population.un.org/wup/Publications/Files/WUP2018-Highlights.pdf).
The other three measurements refer to:

- **city**: city proper, defined by local administrative boundaries
- **metro**: according to UNICEF, the Metro Region is an area "comprising the
urban area as a whole and its primary commuter areas... In addition to the city
proper, a metropolitan area includes both the surrounding territory with urban
levels of residential density and some additional lower-density areas that are
adjacent to and linked to the city." See
[UNICEF](https://www.unicef.org/sowc2012/statistics.php) for more information.
- **urban**: an area that consists of a "continuously built up land mass of
urban development that is within a labor market (metropolitan area or
metropolitan region) and contains no rural land".

Several of the questions in this and the following notebooks concern the
analysis of the relationship between these various definitions.

### Scatterplot

In the code below, write the code to produce a scatter plot with population on
the x-axis and the city population on the y-axis. I have set `warning=FALSE`
because there are some missing data values and it will be helpful to turn off
the warnings about these data rows.

```{r, warning=FALSE}
cities %>%
  ggplot() +
    geom_point(aes(x = population, y = city_pop))
```

Now, in the code region below, produce a scatter plot with population on the
x-axis and metro population on the y-axis.

```{r, warning=FALSE}
cities %>%
  ggplot() +
    geom_point(aes(x = population, y = metro_pop))
```

Which of the two regional populations (metro or city) seem to more closely
relate to the U.N.'s definition of a cities population (e.g., the `population`
variable). **Answer**: The metropolitan population is much closer to the UN
definition than the city population variable.

## Text Plot

Produce a plot with population on the x-axis and metro population on the y-axis
and label each point with the city name.

```{r, warning=FALSE}
cities %>%
  ggplot() +
    geom_text(aes(x = population, y = metro_pop, label = city))
```

What two cities seem to have the largest (absolute) discrepancy between their
UN population and the metro population? **Answer**: Jakarta and Seoul seem to
stand out the most (their metro populations are much larger than the offical
UN population numbers).

Produce a plot with a text geometry where longitude (`lon`) is on the x-axis,
latitude is on the y-axis (`lat`), and the labels are given by the city names
(`name`).

```{r}
cities %>%
  ggplot() +
    geom_text(aes(x = lon, y = lat, label = name))
```

This should roughly resemble a map of the world. In the code below, produce
a modified version of this plot, replacing the text layer with two layers:
a text repel layer and well as a point layer.

```{r}
cities %>%
  ggplot() +
    geom_text_repel(aes(x = lon, y = lat, label = name)) +
    geom_point(aes(x = lon, y = lat))
```

Which of these two plots do you find easier to read, and why? (Note, there is
no clear correct answer here). **Answer**: I slightly prefer the second, but it
would be better to, perhaps, show only a subset of the labels. We will see how
to do this soon!
