---
title: "Notebook 13 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
```

## Movies Data

Over the next few classes we will be working with a dataset of movies I have
constructed consisting of the top 100 grossing films for each year from 1970 to
2019. The data comes from IMDb. Today we will focus on getting familiar with
the various components of the data. Let's read in the four tables of data,
as well as a data dictionary, and then go through each of the tables.

```{r, message=FALSE}
movies <- read_csv(file.path("data", "movies_50_years.csv"))
m_genre <- read_csv(file.path("data", "movies_50_years_genre.csv"))
m_people <- read_csv(file.path("data", "movies_50_years_people.csv"))
m_dict <- read_csv(file.path("data", "movies_50_years_data_dictionary.csv"))
m_color <- read_csv(file.path("data", "movies_50_years_color.csv"))
```

See notebook11 and the data dictionary for more information about the available
variables.

### Movie Joins



```{r}
m_people %>%
  filter(role == "starring") %>%
  filter(rank == 1) %>%
  filter(gender_conf > 0.95) %>%
  left_join(movies, by = c("year", "title")) %>%
  group_by(gender) %>%
  summarize(sm_mean_cl_normal(rating)) %>%
  ggplot(aes(gender, rating_mean)) +
    geom_pointrange(aes(ymin = rating_ci_min, ymax = rating_ci_max)) +
    theme_sm()
```


### Textual Table

```{r}
movies %>%
  mutate(tokens = stri_extract_all(description, regex = "[\\w]+")) %>%
  select(year, title, tokens) %>%
  unnest(cols = c(tokens))
```

```{r}
m_tokens <- movies %>%
  mutate(token = stri_extract_all(description, regex = "[\\w]+")) %>%
  select(year, title, token) %>%
  unnest(cols = c(token)) %>%
  mutate(token = stri_trans_tolower(token))

m_tokens
```

```{r}
m_tokens %>%
  group_by(token) %>%
  mutate(sm_count()) %>%
  group_by(year, title) %>%
  arrange(count) %>%
  slice(1)

movies %>%
  filter(title == "Cliffs of the Wailing River") %>%
  select(year, title, description)
```
