---
title: "Notebook 08A"
output: html_document
---

```{r, include=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
```

## Instructions

This notebook includes several tasks to practice and extend your knowledge of
the material covered within the chapter. For the most part, every notebook with
a name ended in "A" will work with a collection of data about cities and their
population. The questions tend to be relatively contained; the focus is
primarily on writing code to perform a specific task. For this purpose there
will usually be an empty code block for you to put your code. There are also
questions requiring a short response, prompted by the phrase "**Answer:**".
Your answers to these can be write directly in this notebook. Occasionally
new functions or concepts will be shown within each notebook.

Notebooks ending in letters other than "A" more often include new datasets and
open-ended prompts for using the data skills covered in the chapters to address
a wide range of questions. It is suggested to start with the "A" notebook for
each chapter and then trying to work through at least one additional data
project notebook.

When you are done with your code, hitting the *knit* button in the toolbar will
produce an HTML webpage with all of your results. Note that the first time you
run the knit function may require installing some packages. Knitting your code
is a good way to help check that you have written all of the code correctly and
have not accidentally modified anything as you worked through the notebook.

Two important pieces of advice when working through the notebooks:

- It is a best practice to close RStudio everytime that you finish with a
notebook and before you start working on another dataset. This starts each
notebook from scratch and does not risk accidentally using variables created in
one notebook within the scope of another.

- Notice that the notebook file does not automatically word-wrap the way you
may be used to in a word processor. Generally, you should manually hit enter to
start typing on a new line. I try to make sure that both my text and my code are
not very wide. Make sure to do the same with your code and responses. Also, make
sure that your notebook window is wide enough so that my text does not forceable
try to wrap. This is very hard to read and makes it much easier to work through
the notebook.

## Movies data

In this notebook you will create two small datasets, read them into R, and
produce a plot using the skills covered in the first 8 chapters.

For the first task, create a dataset of (at least) ten movies. These can be your
favorite movies, your least favorite movies, or just a random collection that
come to mind. At a minimum, record the following items for each movie:

- title
- date of release (year, month, and day)
- budget
- box-office numbers
- genre
- personal rating of the film, from 1-5 stars

Once you are finished, save the dataset as either a csv file or an excel
worksheet and read it into R in the following block of code:

```{r, message=FALSE}

```

Then, produce a scatterplot below with the budget on the x-axis, box-office
numbers on the y-axis, points colored by the year of release, and the size of
the points corresponding to your personal rating. Use a text repel layer to
label the movies by their title. Include axis labels and titles, along with
units of measurement.

```{r}

```

If you think it is needed, also include logarithmic scales for the x- and
y-axes.

## Friends data

For the second dataset, you are going to collect information about ten of your
friends. At a minimum, record the following variables:

- their name
- where you met them
- the date you met them (year, month, day; approximate if not sure)
- where they grew up (can use city, country, state; use whatever makes sense for
  your data)
- personal rating of the friend, from 1-5 stars

Once you are finished, save the dataset as either a csv file or an excel
worksheet and read it into R in the following block of code:

```{r, message=FALSE}

```

Then, produce a scatter plot below using any two variables that you find
interesting as the x- and y-axes. Include a highlight layer showing a subset
of your friends. These could be friends you met at a certain time in your life,
highlighting your best friend(s) or something else that is interesting to you.
Include axis labels and titles, along with units of measurement.

```{r}

```
