---
title: "Notebook 01A"
output: html_document
---

```{r, include=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
```

## Instructions

This notebook includes several tasks to practice and extend your knowledge of
the material covered within the chapter. For the most part, every notebook with
a name ended in "A" will work with a collection of data about cities and their
population. The questions tend to be relatively contained; the focus is
primarily on writing code to perform a specific task. For this purpose there
will usually be an empty code block for you to put your code. There are also
questions requiring a short response, prompted by the phrase "**Answer:**".
Your answers to these can be write directly in this notebook. Occasionally
new functions or concepts will be shown within each notebook.

Notebooks ending in letters other than "A" more often include new datasets and
open-ended prompts for using the data skills covered in the chapters to address
a wide range of questions. It is suggested to start with the "A" notebook for
each chapter and then trying to work through at least one additional data
project notebook.

When you are done with your code, hitting the *knit* button in the toolbar will
produce an HTML webpage with all of your results. Note that the first time you
run the knit function may require installing some packages. Knitting your code
is a good way to help check that you have written all of the code correctly and
have not accidentally modified anything as you worked through the notebook.

Two important pieces of advice when working through the notebooks:

- It is a best practice to close RStudio everytime that you finish with a
notebook and before you start working on another dataset. This starts each
notebook from scratch and does not risk accidentally using variables created in
one notebook within the scope of another.

- Notice that the notebook file does not automatically word-wrap the way you
may be used to in a word processor. Generally, you should manually hit enter to
start typing on a new line. I try to make sure that both my text and my code are
not very wide. Make sure to do the same with your code and responses. Also, make
sure that your notebook window is wide enough so that my text does not forceable
try to wrap. This is very hard to read and makes it much easier to work through
the notebook.

## Running R Code

In the code block below, divide the number 57 by 3.

```{r}
57 / 3
```

Is the number 57 a prime number? **Answer**:

In the code block below, make a variable named `fav_number` and set it equal to
your favorite number. (Note, the quotes are not part of the variable name).

```{r}

```

In the code below, apply the function `log10` to your favorite number. (Again, the
quotes are not part of the function name).

```{r}

```

Note that the only point of the exercises here are to familiarize you with running
code in R, creating variables, and applying functions. We will get to more
interesting  tasks in the next chapter.

## Running Functions

R contains several functions for producing (pseudo)-random variables. These are
useful, for example, when using R to run simulation models. For example the
function `runif` selects (by default) a set of random numbers between 0 and 1.
It takes one required arugment named `n`, which indicates how many random numbers
should be generated. In the code below, use the `runif` function to produce 100
random numbers. Verify that each time you run the code a different set of 
numbers is produced:

```{r}

```

The `runif` function also has two optional parameters. These are named `min` and
`max`; they determine the lower and upper bounds from which random numbers should
be generated. By default these are set to 0 and 1. In the code below, generate 100
random numbers between 50 and 100. Here, select the number of random numbers
*by position* (that is, without the code `n = 100`).

```{r}

```

A common concern for new R users when running the code above is the meaning behind
the numbers in square brackets to the left of the output. These are not part of the
output itself. Instead, they provide a counter telling you which result number appears
just to the left of right of it. The first line will always start with `[1]` because 
it always starts with the first result. The second line's number will depend on the
width of your screen when you ran the code.

## Largest Cities Dataset

We will make extensive use of a dataset that has information about the largest
cities (by population) in the world. To read in and print out this dataset, run
the following lines of code:

```{r, message=FALSE}
cities <- read_csv(file.path("..", "data", "largest_cities.csv"))
cities
```

Looking at the data, try to answer the following questions.

How many rows are in the data set? **Answer**:

What are the observations in the dataset. **Answer**:

List three of the variables in the dataset. **Answer**:

The population variables are given in millions of people. How many
people live in all of Tokyo? **Answer**:

How people live in Atlanta (Note: you will have to open the data
explorer to see this data)? **Answer**:

## Formating Code

In the code block below, I wrote some code to add a new column to the dataset
that describes the population density for each city (1000s of people per square
kilometer) and sorts from the most dense to the least dense. We will learn
this code over the next 6 chapters. For now, I want to focus on formatting
the code correctly. I did not include any spaces! Put the correct spaces into
the code to make it match the style guide given in the notes.

```{r}
new_data<-mutate(cities,city_density=city_pop/city_area*1000)
new_data<-arrange(new_data,desc(city_density))
select(new_data,name,country,city_density)
```

Make sure to run the code after you are done. What city in the dataset has, on
average, the most people per square kilometer? **Answer**:
