---
title: "Notebook 12 -- Solutions"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
```

## Movies Data

Over the next few classes we will be working with a dataset of movies I have
constructed consisting of the top 100 gross films for each year from 1970 to
2019. The data comes from IMDb. Today we will focus on getting familiar with
the various components of the data. Let's read in the four tables of data,
as well as a data dictionary, and then go through each of the tables.

```{r, message=FALSE}
movies <- read_csv(file.path("data", "movies_50_years.csv"))
m_genre <- read_csv(file.path("data", "movies_50_years_genre.csv"))
m_people <- read_csv(file.path("data", "movies_50_years_people.csv"))
m_dict <- read_csv(file.path("data", "movies_50_years_data_dictionary.csv"))
m_color <- read_csv(file.path("data", "movies_50_years_color.csv"))
```


### Movies Table Subsets

Here, we are going to explore a few different ways to subset the movies dataset.
For some analyses it is possible to use the entire dataset, but for others it
may help to reduce the size and scope a bit. We will focus on the types of
subsetting that are a bit less obvious. You do not need to save the subsets of
data; just plot or print them to the notebook as needed.

To start, select the 10 movies from each year that have the highest number of
IMDb ratings:

```{r, question-08}
movies %>%
  arrange(desc(rating_count)) %>%
  group_by(year) %>%
  slice(1:10)
```

Next, select every movie that made more than 10% (gross) of the maximum gross
budget for a film in the year associated with the film. Count the number of
films selected in each year by this metric and show a bar plot of the pattern
over time.

```{r, question-09}
movies %>%
  group_by(year) %>%
  mutate(sm_max(gross)) %>%
  filter(gross > 0.1 * gross_max) %>%
  summarize(sm_count()) %>%
  ggplot(aes(year, count)) +
    geom_col()
```

We can also select a specific sheet of movies based on a search string using
the `stri_detect` function, like this:

```{r, question-10}
movies %>%
  filter(stri_detect(title, fixed = "Lord of the Rings:"))
```

Now, use a similar approach to extract all of the Harry Potter movies in the
dataset:

```{r, question-11}
movies %>%
  filter(stri_detect(title, fixed = "Harry Potter"))
```

You can do a similar search for terms in the description field. Do this
below to find instances of the term "alien":

```{r, question-12}
movies %>%
  filter(stri_detect(description, fixed = "alien"))
```

One issue with the code above is that it will not find terms where alien is
captialized. The function `stri_trans_tolower` takes a string and returns a
non-captilized version of it. In the code below, create a lowered version of
the description variable and detect "alien" again; did this add any new films?

```{r, question-13}
movies %>%
  mutate(description_lower = stri_trans_tolower(description)) %>%
  filter(stri_detect(description_lower, regex = "alien"))
```

Using the code you created above, draw a bar plot showing the number of
alien films in each year:

```{r, question-14}
movies %>%
  mutate(description_lower = stri_trans_tolower(description)) %>%
  filter(stri_detect(description_lower, regex = "alien")) %>%
  group_by(year) %>%
  summarize(sm_count()) %>%
  ggplot(aes(year, count)) +
    geom_col()
```

Finally, we can use a similar approach to create a new variable rather than
filtering the data. Below, create a variable named `term_alien` based on
whether a film has the term alien in its description. Compute a confidence
interval plot showing the distribution of the variable `poster_brightness`.

```{r, question-15}
movies %>%
  mutate(description_lower = stri_trans_tolower(description)) %>%
  mutate(term_alien = stri_detect(description_lower, regex = "alien")) %>%
  group_by(term_alien) %>%
  summarize(sm_mean_cl_normal(poster_brightness)) %>%
  ggplot(aes(term_alien, poster_brightness_mean)) +
    geom_pointrange(aes(ymin = poster_brightness_ci_min, ymax = poster_brightness_ci_max))
```

Take note of the pattern that you see. We will see how to automate this process of
detecting terms in free text fields in a future class.
