---
title: "Project 03"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(lubridate)
library(sf)
library(units)
library(RcppRoll)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
options(lubridate.week.start = 1)
Sys.setlocale(locale = "en_US.UTF-8")

sm_centroid <- function(data) {
  suppressWarnings({ z <- st_coordinates(st_centroid(data)) })
  return(tibble(lon = z[,1], lat = z[,2]))
}
```

## US County COVID-19

This project is more like a graded notebook than an independent
project. There are several questions that you need to address with
the data. Once you are finished knit the document and hand in the
HTML file with your answers.

Unless otherwise noted, you do not need to include manual labels
in the answers.

### Load the Data

There are three datasets that you will use in this project. They
are similar to the datasets we had with the French data: a spatial
object (county), a dataset of demographic data, and a dataset of
daily Coronavirus records at the county level.

```{r, message = FALSE}
county <- read_sf(file.path("data", "us_counties.geojson"))
covid <- read_csv(file.path("data", "us_county_covid.csv"))
demog <- read_csv(file.path("data", "us_county_metadata.csv"))
```

The COVID-19 dataset is a bit simplier than the one from France. It
has only new case counts and newly reported deaths. Note that these
are only the counties with complete data from 01 April 2020.

```{r}
covid
```

The demographic data gives several points of information about each county. Unless
otherwise noted, the data comes from 2019.

```{r}
demog
```

Most of the variables should be self-explanatory. Unemployment rate is in percentages (0-100),
median household income in USD. February and August average temperatures are in Celsius. The
`physicians_per_pop` give the number of active physicians per 100k residents according to the
AAMC in 2018. Note that there is information for almost every county in the US, even if there
are not records in the COVID-19 dataset.

### Showing Spatial Data

Plot the average February temperature in each county, excluding Alaska (AK), Hawaii (HI) and
Puerto Rico (PR). Use `scale_fill_distiller` with an appropriate color scale and an appropriate
spatial projection.

```{r}
county %>%
  inner_join(demog, by = c("fips", "state")) %>%
  filter(!(state %in% c("AK", "HI", "PR"))) %>%
  st_transform(5070) %>%
  ggplot() +
    geom_sf(aes(fill = feb_temp_avg), size = 0) +
    scale_fill_distiller(palette = "Spectral", guide = "legend", n.breaks = 10) +
    theme_void()
```

Now, show the unemployment rate for each county in Puerto Rico (it's one of the few
variables we have access to). Make sure to choose a reasonable projection
(seach http://epsg.io/ perhaps?) and color scale.

```{r}
county %>%
  inner_join(demog, by = c("fips", "state")) %>%
  filter(state == "PR") %>%
  st_transform(3991) %>%
  ggplot() +
    geom_sf(aes(fill = unemployment_rate), size = 0) +
    scale_fill_distiller(palette = "Spectral", guide = "legend", n.breaks = 10) +
    theme_void()
```

Finally, show the number of new coronavirus deaths on 1 May 2020 for each county, excluding
Alaska (AK), Hawaii (HI) and Puerto Rico (PR). Use an appropriate color scale and spatial
projection.

```{r}
county %>%
  inner_join(demog, by = c("fips", "state")) %>%
  inner_join(filter(covid, date %in% ymd(c("2020-05-01")))) %>%
  filter(!(state %in% c("AK", "HI", "PR"))) %>%
  st_transform(5070) %>%
  ggplot() +
    geom_sf(aes(fill = cases), size = 0) +
    scale_fill_distiller(palette = "Spectral", guide = "legend", n.breaks = 10) +
    theme_void() +
    facet_wrap(~date)
```

What three regions seem to have the worst numbers as of 1 May? **Answer**:

### Time Series Data

Create a dataset of the 8 counties with the largest population:

```{r}
top_counties <- arrange(demog, desc(population)) %>%
  slice(1:8)
```

Now, show a time series plot with time on the x-axis and number of cases on the y-axis
for the 8 counties with the largest population. Color the lines based on the county
and force the x-axis to label each month in the data with the month name.

```{r}
covid %>%
  semi_join(top_counties, by = c("fips", "state")) %>%
  ggplot(aes(date, cases)) +
    geom_line(aes(color = county)) +
    scale_x_date(date_breaks = "month", date_labels = "%B")
```

For each county, compute the number of new cases per 100k residents for each
day. Then, compute the average rate (new cases per 100k people) of new cases
as a function of the day of the week. Show the relationship between day of the
week and the rate mean on a scatterplot.

```{r}
covid %>%
  inner_join(demog, by = c("fips", "state")) %>%
  mutate(rate = cases / population * 100000) %>%
  mutate(wday = wday(date, label = TRUE, abbr = FALSE)) %>%
  group_by(wday) %>%
  summarize(sm_mean(rate)) %>%
  ggplot(aes(wday, rate_mean)) +
    geom_point()
```

### Window Functions

```{r}
covid_lag <- covid %>%
  group_by(fips, state) %>%
  mutate(cases_1_week = lag(cases, n = 7), cases_2_week = lag(cases, n = 2), cases_3_week = lag(cases, n = 14))
```


```{r}
covid_lag %>%
  ungroup() %>%
  summarize(sm_cor(deaths, cases), sm_cor(deaths, cases_1_week), sm_cor(deaths, cases_2_week), sm_cor(deaths, cases_3_week))
```


```{r}
covid_lag <- covid %>%
  mutate(cases = roll_meanr(cases, n = 7), deaths = roll_meanr(deaths, n = 7)) %>%
  group_by(fips, state) %>%
  mutate(cases_1_week = lag(cases, n = 7), cases_2_week = lag(cases, n = 14), cases_3_week = lag(cases, n = 21))
```


```{r}
covid_lag %>%
  ungroup() %>%
  summarize(sm_cor(deaths, cases), sm_cor(deaths, cases_1_week), sm_cor(deaths, cases_2_week), sm_cor(deaths, cases_3_week))
```

### Demographics

```{r}
covid %>%
  group_by(fips) %>%
  summarise(cases = sum(cases), deaths = sum(deaths)) %>%
  left_join(demog, by = "fips") %>%
  mutate(party = if_else(per_dem_2016 > 0.5, "Democrate", "Republican")) %>%
  group_by(party) %>%
  summarise(rate = sum(deaths) / sum(cases))
```

```{r}
covid %>%
  filter(date == "2020-10-01") %>%
  group_by(fips) %>%
  left_join(demog, by = "fips") %>%
  mutate(party = if_else(per_dem_2016 > per_gop_2016, "D", "R")) %>%
  group_by(party) %>%
  summarise(rate = sum(deaths))
```

```{r}
covid %>%
  filter(date == "2020-10-01") %>%
  summarise(sum(deaths))
  group_by(fips) %>%
  left_join(demog, by = "fips") %>%
  ungroup() %>%
  mutate(income = if_else(median_hh_income > mean(median_hh_income), "rich", "poor")) %>%
  group_by(income) %>%
  summarise(rate = sum(deaths))
  

covid %>%

```


```{r}
covid %>%
  filter(date == "2020-10-01") %>%
  left_join(demog, by = c("fips", "state")) %>%
  group_by(state) %>%
  summarise(rate = sum(deaths) / sum(cases), dem_mean = weighted.mean(per_dem_2016, population)) %>%
  #summarise(sm_cor(rate, dem_mean))  
  ggplot(aes(dem_mean, rate)) +
    geom_text(aes(label = state))
```






