\documentclass[12pt, a4paper]{article}

\usepackage{fontspec}
\usepackage{geometry}
\usepackage{lastpage}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}
\usepackage[normalem]{ulem}
\usepackage{soul}
\usepackage{multicol}
\usepackage[dvipsnames]{xcolor}

\geometry{
  top=2cm,
  bottom=3cm,
  left=2cm,
  right=2cm,
  marginparsep=4pt,
  marginparwidth=2cm
}

\definecolor{lgrey}{rgb}{0.8, 0.8, 0.8}
\definecolor{solarized@magenta}{HTML}{D33682}
\newcommand{\magenta}[1]{\textcolor{solarized@magenta}{#1}}

\renewcommand{\headrulewidth}{0pt}
\pagestyle{fancyplain}
\fancyhf{}
\lfoot{\color{lgrey}2020-09-01}
% \rfoot{\color{lgrey}page \thepage\ of \pageref{LastPage}}

\setlength{\parindent}{0pt}
\setlength{\parskip}{0pt}

\usepackage{xunicode}
\defaultfontfeatures{Mapping=tex-text}

\setromanfont{YaleNew}

\begin{document}

\begin{center}
\textbf{Introduction to Data Science: Participation Grades}
\end{center}

\vspace{0.5cm}

\textbf{Overview:} \vspace{6pt}

A major component of your grade for this semester (30\%) measures your
participation in the course meetings. The goal is to hold everyone
accountable to being present and engaged in our all too short and infrequent
class meetings. You may be familiar with participation grades from other
courses that are primarily discussion based. These often focus on the
frequency and quality of public comments made during class. Participation in
\textit{Introduction to Data Science} works differently. While questions and
comments are encouraged, there simply is not enough time or space to have
everyone share their work within a typical class meeting. \\

As an alternative model, students are asked to fill out a short form (linked
to on the course website) at the end of each class meeting. In addition to
two short questions, space is included in the form to include a free
response question. During class, one or more elements from
the day's notebook will be flagged to be uploaded to this form. This may,
for example, be a block of code for a specific question or a response to a
short-answer question. The form must be filled out immediately at the end of
class, and the assignment may be different between the two course sections. \\

General attendance, including arriving on time and being ready to share your
screen or answer a question when called on, are also important elements of
the participation grade.

\bigskip

\textbf{Rubric:} \vspace{6pt}

Below are descriptions of what is expected for a participation grade of an
\textbf{\magenta{A}} and a grade of \textbf{\magenta{C}}. Other intermediate
grades will be given for students who fall somewhere in-between these two
categories. Failing to fulfill the requirements of a C will be result in an
\textbf{\magenta{F}} for participation. \\

\setlength{\leftskip}{1cm}

\textbf{Grade of \magenta{A}}: The material submitted in the class forms
is typically correct and answers the questions as posed in the notebook. For
questions of an open-ended nature, the solutions illustrate a desire to think
critically and creatively about the data and material. For this grade, a student
should rarely be late, miss class, be unprepared
to answer a question or share their screen, or fail to submit a class form.
Generally expected to miss no more than two class meetings or class forms.
\\

\textbf{Grade of \magenta{C}}: Submitted lab forms generally show an attempt to
complete the assigned questions. Behaviors such as arriving late, missing class,
failing to submit a form, or being unprepared to screen share or answer a
question should not be habitual. Generally expected to miss no more than four
class meetings or class forms.

\setlength{\leftskip}{0pt}

\bigskip

\textbf{Note on Absences:} \vspace{6pt}

As described above, all students are allowed two ``free'' absences or failures
to submit a form before it will impact a participation grade. These are usually
sufficient to accommodate most unexpected circumstances, such as computer issues,
illness, or one-off unavoidable conflicts. There is no need to email about or
explain these first two class absences. If a situation arises where you need to
miss additional classes, please contact the instructor as soon as possible to
find a suitable solution in order for it to not impact your participation
grade.\\

Students are welcome to attend either course section.

\end{document}
