\documentclass[12pt, a4paper]{article}

\usepackage{fontspec}
\usepackage{geometry}
\usepackage{lastpage}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}
\usepackage[normalem]{ulem}
\usepackage{soul}
\usepackage{multicol}
\usepackage[dvipsnames]{xcolor}

\geometry{
  top=1cm,
  bottom=2cm,
  left=1cm,
  right=1cm,
  marginparsep=4pt,
  marginparwidth=1cm
}

\definecolor{lgrey}{rgb}{0.8, 0.8, 0.8}

\renewcommand{\headrulewidth}{0pt}
\pagestyle{fancyplain}
\fancyhf{}
\lfoot{\color{lgrey}2020-08-14}
% \rfoot{\color{lgrey}page \thepage\ of \pageref{LastPage}}

\setlength{\parindent}{0pt}
\setlength{\parskip}{0pt}

\usepackage{xunicode}
\defaultfontfeatures{Mapping=tex-text}

\setromanfont{YaleNew}

\begin{document}

\begin{center}
\textbf{MATH/STAT 289: Introduction to Data Science, Fall 2020}
\end{center}

\noindent
\begin{tabular}{ l l }
\textbf{Instructor:} &  \textbf{Taylor Arnold} \\
E-mail: & \texttt{tarnold2@richmond.edu} \\
Website: & \texttt{https://statsmaths.github.io/stat289-f20}
\end{tabular}

\vspace{0.5cm}

\textbf{Description:} \vspace{6pt}

Data science is an interdisciplinary field concerned with drawing knowledge
from data and communicating those results to various audiences. Data
science needs to be learned \textit{by doing} data science. At the end of the semester,
students will have acquired a toolkit of methods, and the knowledge of how to
use them in practice, to address important social, cultural, and scientific
questions with data-driven techniques.

\bigskip

\textbf{Prerequisites:} \vspace{6pt}

The pace of this course assumes that students have had some prior exposure to a programming
language and have taken a course in which statistical techniques were applied to the
analysis of real-world datasets. Prior knowledge of R is useful but not required. As an
alternative starting point for anyone without this background, we suggest students consider
taking MATH209. Several sections of this course are being offered this semester.

\bigskip

\textbf{Website and Software:} \vspace{6pt}

All of the materials and assignments for the course will be posted online on the course's
website. These notes will continue to be available for your reference after the semester has
finished. This course will make use on an online platform called RStudio Cloud. Students
will be asked to sign-up for an account (internet access and a modern browser are required;
no other software is needed).

\bigskip

\textbf{Method of Instruction, Fall 2020:} \vspace{6pt}

Due to social distancing guidelines and the need to maximize flexibility, Data Science is
being taught using online instruction. Classes will be presented remotely over Zoom.
Information about office hours and other logistics will be presented at the first course
meeting. In the event of changes to policies regarding residence and instruction, the
course structure and grading scheme of the course should remain the same; all reasonable
requests for deadline extensions and accommodations will be honored.

\bigskip

\textbf{Grading:} \vspace{6pt}

There will be several class projects assigned throughout the semester. Each project will be
distributed with a rubric and assigned a total number of overall points. Projects will be
submitted through Box. At the end of the semester, the overall project grade will be
determined by adding together the total number or earned points and dividing by the total
number of available points. Final grades will be determined according the following:

\begin{itemize}\setlength\itemsep{0em}
  \item \textbf{Projects:} 70\%
  \item \textbf{Participation:} 30\%
\end{itemize}

Participation grades will be given according to the rubric posted on the website.
There are no exams for this course.

\bigskip

\textbf{Alternative Grading:} \vspace{6pt}

Some students may be unable to travel to campus due to the ongoing COVID-19 pandemic. It is
the University's policy that all enrolled students should be prepared to attend courses
remotely from wherever they are; in reality this may be difficult due to shifted
time-zones and situations at home. As an alternative, students will be given the option to
follow the course material asychronously and receive a final grade based entirely on their
project scores. In the interest of equity, this option is open to all students, but must be
chosen during the first week of the semester.

\bigskip

\textbf{Additional Policies:} \vspace{6pt}

% \begin{itemize}\setlength\itemsep{0em}
% \item \textbf{Academic honesty:}
% Cheating and plagiarism are grave scholarly offenses and potential grounds
% for expulsion; they are also a major barrier to your intellectual development.
% You are expected to familiarize yourself with the entirety of the
% University of Richmond’s Honor Code.
% \item \textbf{Pass/Fail/Withdraw/Incomplete:} If you choose to take this
% class pass/fail, it is expected that you will achieve a minimum grade of a B- (my project
% rubrics do not specify grades lower than this). I am generally willing to allow withdraws
% after the deadline, with approval of your Dean, without penalty. However, I typically only
% allow a grade of incomplete (I or Y) in cases where at least half of the semester's projects
% have been completed.
% \end{itemize}

A more detailed introduction to the class and its contents are given in the notes provided
on the course website. If you have any questions regarding the course policies that are not
covere, or that you find unclear, please ask for clarification at any point in the semester.

\end{document}
