---
title: "04. Organizing Data"
author: ""
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r, include=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)

food <- read_csv(file.path("data", "food.csv"))
food_prices <- read_csv(file.path("data", "food_prices.csv"))
```

## Verbs

In this chapter we are going to cover a set of functions that take
a data frame as an input and return a new version of the data frame. These
functions are called *verbs* and come from the **dplyr** package. If you are
familiar with running database queries, note that all of these verbs
map onto SQL commands. In fact, R can be set up so that
**dplyr** is called over a database rather than a local
data frame in memory.

There are over 40 verbs in the **dplyr** package, though most are
a minor variant or specific application of another verb. In this chapter
we will see only four of them, all of which are related to selecting and
arranging rows and columns:

- select a subset of rows from the original dataset (`filter` and `slice`)
- select a subset of columns from the original dataset (`select`)
- sort the rows of a dataset (`arrange`)

In all verb functions, the first argument is the original data frame and the
output is a new data frame. Here, we will also see the functions `between` and
`%in%` to assist with the filtering command and `desc` to assist with arranging
the rows of a dataset.

Note that **verbs do not modify the original data**; they operate on a copy of
the original data. We have to make an explicit name for the new dataset if we
want to save it for use elsewhere.

## Choosing rows

It is often useful to take a subset of the rows of an existing dataset, for
example if you want to build a model on a certain subpopulation or highlight a
particular part of the data in a plot. Perhaps the most straightforward way to
take a subset of rows is to indicate the specific row numbers that we want to
extract. In order to select rows by row numbers, we use the verb `slice`, followed
by the numbers of the rows we want seperated by commas. Here is an example taking
the second, fifth, and seventh rows of the data:

```{r}
food %>%
  slice(2, 5, 7)
```

As mentioned above, the code here does not change the dataset `food` itself. It
still has all 61 rows of food contained in it. If we want to create a new dataset
with just these three food item, we need to explicitely name and assign it. For
example, here is how we would create a dataset of the first five food items
named `food_first_five`:

```{r}
food_first_five <- food %>%
  slice(1, 2, 3, 4, 5)
```

There a convenient a shorthand for selecting a range of row numbers, for example
every row from the tenth to the twentieth, by indicating the starting and ending row
number by a colon. Here, for example, is another way to select the first five rows of
the dataset:

```{r}
food %>%
  slice(1:5)
```

Another way to take a subset of our data is to select rows based on conditions
about the variables in the dataset. To do this we use the `filter` function,
which accepts a statement about variable in the dataset. Only rows where the
statements are true will be returned. For example, here is how we use the
`filter` command to select the foods that have more than 150 calories grams of
sugar in each serving:

```{r}
food %>%
  filter(calories > 150)
```

The output dataset has only 20 rows, compared to the 62 in the original data.
Other comparisons can be done with `<`, `>=` and `<=`. There is also a special
function called `between` that is often useful. For example, here are the rows
that have between 2 and 3 grams of total fat:

```{r}
food %>%
  filter(between(total_fat, 2, 3))
```

If you want to filter on a categorical variable, you can use the `%in%`
operator to select specific categories. Here is the code to filter only the
fish and vegetable variables:

```{r}
food %>%
  filter(food_group %in% c("fish", "vegetable"))
```

As with the other verbs, we can chain together multiple calls to produce more
complex logic. For example, this code selects fruits that have more than 150
calories per serving:

```{r}
food %>%
  filter(calories > 150) %>%
  filter(food_group %in% c("fruit"))
```

Which results in a reduced dataset with only 1 row (avocados).

## Data and Layers

Now that we know how to create a subset of our data, let's use this new knowledge to
build some interesting data visualizations. To start, create a dataset that just
consists of the food types that are in the meat food group:

```{r}
food_meat <- filter(food, food_group %in% c("meat"))
food_meat
```

You should know from the previous chapter how to create a scatterplot of `calories` and `total_fat`
with just the meat data. Here is the code:

```{r}
food_meat %>%
  ggplot() +
    geom_point(aes(x = calories, y = total_fat))
```

One of the core ideas behind the Grammar of Graphics is that complex visualizations can be
constructed by layering relatively simply elements on top of one another. What if we wanted to
put together two layers where one layer uses the `food` dataset and the other uses `food_meat`?
To do this, we can *override* the default dataset in a layer with the option `data =`. This will
use a different dataset within a particular layer. For example, here is how we can layer the
meat dataset on top of the rest of the food items.

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = total_fat)) +
    geom_point(aes(x = calories, y = total_fat), data = food_meat)
```

This plot, however, does not look any different than it would if we were just to plot all of the
food together. The second layer of points just sits unassumingly on top of the rest of the data.
To rectify this, we can color each layer a different color in order to distinguish them from one
another. Let's try to highlight the meat food group in a navy blue, while making the rest of the
points a light grey:

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = total_fat), color = "grey85") +
    geom_point(aes(x = calories, y = total_fat), color = "navy", data = food_meat)
```

We now have a plot that shows exactly where the meats are relative to the other food items. We can
further build up the plot by showing the names of just these rows of the dataset as well:

```{r}
food %>%
  ggplot() +
    geom_point(aes(x = calories, y = total_fat), color = "grey85") +
    geom_point(aes(x = calories, y = total_fat), color = "navy", data = food_meat) +
    geom_text_repel(
      aes(x = calories, y = total_fat, label = item),
      color = "navy",
      data = food_meat
    )
```

Notice how a relatively small set of commands can be put together in different ways to build
a variety of plots. Already, we are making further progress towards building informative and
beautiful graphics in R!

## Selecting Columns

It is also possible to take a subset of the columns in a dataset. To do this, we make use of
the verb `select`. We pass it the names of the variables we want to keep in the output dataset,
in the (possibly new) order that we want the columns to be arranged in. Here, for example, is
a new version of the foods dataset containing only the food item name followed by the amount
of Vitamin A and Vitamin C:

```{r}
food %>%
  select(item, vitamin_a, vitamin_c)
```

We will not need to use the `select` verb as often as `filter` because for the most part
having extra variables around does not effect data visualizations or data models. However,
it can be useful to displaying results. As we saw above, the Vitamin A and Vitmain C columns
were cut-off in the original output but are not visable in the selected dataset version.
Removing and reordering unneeded columns will also be useful in some of the advanced
applications that are discussed in the following chapters.

## Arranging Rows

The verbs `slice` and `filter` determine a subset of rows to keep from the original dataset.
The `arrange` verb, in contrast, keeps all of the original data but re-orders its rows.
Specifically, we give it one or more variable names and it sorts the data by the first variable
from smallest to largest (or alphabetically for character variables). In the case of ties,
the second variable is used if given. More variables can be given to further break additional
ties. Here is an example where we order the dataset first by `food_group` and then by `calories`:

```{r}
food %>%
  arrange(food_group, calories)
```

In the new dataset all of the dairy products come up first followed by the fish products.
Within each group, the items are sorted from the lowest to highest number of calories.

The ordering can be reversed (i.e., from the highest to the lowest value) be wrapping a
variable in the function `desc()`, such as this ordering from the most saturated fat to
the least:

```{r}
food %>%
  arrange(desc(sat_fat))
```

In the result here, "Cheese" has been placed at the top of the dataset, followed
by "Sour Cream" and "Lamb".

One particularly useful application of `arrange` is to pair it with the verb `slice`.
Here, for example, is the code to select the six foods in our dataset that have the
highest amount of Vitamin A:

```{r}
food %>%
  arrange(desc(vitamin_a)) %>%
  slice(1:6)
```

By saving this dataset, we could highlight these specific foods on top of a plot
comparing them to the remainer of the data.

## Exercises and References

There is are two exercise notebook for this chapter:

- **exercies04A.Rmd**: practice applying the four data verbs presented here
- **exercies04B.Rmd**: open-ended questions to explore with a new dataset

It is strongly recommended to complete at least the first notebook prior to
continuing on to the next chapter.
