---
title: "06. Creating New Variables"
author: ""
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

```{r, include=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)

food <- read_csv(file.path("data", "food.csv"))
food_prices <- read_csv(file.path("data", "food_prices.csv"))
```

## Mutate verb

The final core **dplyr** verb that we will look at is used to create a new
variable in our dataset based on other variables that are already present.
This verb is called `mutate`, and works by giving it the name of the variable you
want to create followed by the code that describes how to construct the variable
in terms of the rest of the data.

As an example, consider computing the number of calories in an 200g portion of each
food. All of the variables in the dataset are currently given as 100g portions, so
to compute this we need to multiply the `calories` variables by 2. To do this, we
use the `mutate` verb to name and describe a new variable `calories_200g`.

```{r}
food %>%
  mutate(calories_200g = calories * 2)
```

Notice that there is a new variable named `calories_200g` that has been added as the
last column in the dataset. Because it is added at the end of the dataset, it gets
hidden in the output shown above. Making use of `select` allows us to see the new
values:

```{r}
food %>%
  mutate(calories_200g = calories * 2) %>%
  select(item, food_group, calories, calories_200g)
```

And now we can see that the new column has been created by doubling the number given the
`calories` column.

Note that `mutate` can also be used to modify any existing column in the dataset by
using the name of an extant variable. In this case the position of the variable within the
tables does not change.

The mutate verb itself has a relatively straightforward syntax. The main challenge is
knowing how to apply and chain together the various transformations that are useful
within an analysis. In the next few sections, we highlight several common types of
operations that we will be useful in subsequent applications.

## Conditional values

Many of the uses for the mutate verb involve assigning one value when a set of conditions
is true and another if the conditions are false. For example, consider creating a new variable
called `sugar_level` based on the relative amount of sugar in each food item. We might classify
a food has having a "high" sugar level if has more than 10g of sugar per 100g serving, and a
"normal" amount otherwise. In order to create this variable, we need the function `if_else`.

The `if_else` function has three parts: a TRUE/FALSE statement, the value to use when the
statement is true, and the value to use when it is false. Here is an example to create our
new variable:

```{r}
food %>%
  mutate(sugar_level = if_else(sugar > 10, "high", "normal")) %>%
  select(item, food_group, sugar, sugar_level)
```

Looking at the first rows of data, we see that apples and bananas are classified
as high sugar foods, whereas the other sugar levels are given the sugar level
category of "normal".

The `if_else` function can be used to produce any number of categories by using it
multiple times. Let's modify our sugar level variable to now have three categories:
"high" (over 10g), "low" (less than 1g), and "normal" (between 1g and 10g). There
are several different ways to get to the same result, but I find the easiest is to
start by assigning a default value and then changing the value of the new variable
in sequence. For example, here some code that produces our new categories:

```{r}
food %>%
  mutate(sugar_level = "default") %>%
  mutate(sugar_level = if_else(sugar < 1, "low", sugar_level)) %>%
  mutate(sugar_level = if_else(sugar > 10, "high", sugar_level)) %>%
  mutate(sugar_level = if_else(between(sugar, 1, 10), "normal", sugar_level)) %>%
  select(item, food_group, sugar, sugar_level)
```

In each `if_else` step we are telling the mutate function that if the condition is
false set `sugar_level` equal to itself. In other words, if the condition does not
hold, do not change the value of the variable.

In may wonder why we created a "default" value for the variable `sugar_level`. It
would have been one less line of code to set the default value to "normal" and remove
the final mutate function. The reason for the approach above is three-fold. First,
it's easier to understand what the code is doing in it's current format because each
condition ("high", "normal", and "low") is explicitly coded. Secondly, it creates a
nice check on our code and data. If we find a row of the output that still has the value
"default" we will know that there is a problem somewhere. Finally, the code above
will more safely handle the issues with missing values, and issue that we will
return to shortly.

## Creating labels

Another common type of manipulation that is used with the mutate verb is the
creation of a textual label. This can be done with the function `stri_paste`. It
takes any number of inputs and combines them into a single string. We can mix both
variables in the dataset as well as fixed strings to create useful labels for data
visualiations. For example, if we create a label variable describing each fruit and
its sugar content:

```{r}
food %>%
  mutate(label = stri_paste(item, " (", sugar, "g)")) %>%
  select(item, sugar, label)
```

The `label` variable could then be used as a textual label in a subsequent plot with
textual data. This is similar to the `sm_paste` function that we saw when summarizing
the values of a character vector.

## Factors

R has a special data type called a "factor" (abbreviated "fct") that is specifically
designed to handle categorical variables. It is typically not a good idea to store
data as a factor because the resulting variables have some odd, error-producing,
behaviors. However, it can be useful to create a factor as part of a mutate function
just prior to creating a data visualizations.

For us, biggest difference between factors and character vectors is that a factor
vector has a default ordered of its unique values, called the factor's "levels".
Creating and understanding factors is useful because it allows us to change the
ordering of categories within visualizations and models (which by default is done
alphabetically).

One of the easiest ways to produce a factor variable with a given order is through
the function `fct_inorder`. It will order the categories in the same order that
they (first) appear in the dataset. Combining this with the `arrange` function provides
a lot of control over how categories become ordered. For example, the following code
produces a bar plot of the food groups in our dataset arranged from the largest
category to the smallest category:

```{r}
food %>%
  group_by(food_group) %>%
  summarize(sm_count()) %>%
  arrange(desc(count)) %>%
  mutate(food_group = fct_inorder(food_group)) %>%
  ggplot() +
    geom_col(aes(food_group, count))
```

Other useful functions for manipulating categories include `fct_relevel` for manually putting
one category first and `fct_lump_n` for combining together the smallest categories into a
collective "Other" category.

## Mutate summaries

All of summary functions that were introduced in the previous chapter can also
be applied within the mutate version. Instead of reducing the data to a single
summary row, summarizing within the mutate verb duplicates the summary statistic
in each row of the dataset. Here is an example of including the average number
of calories across all rows of the dataset:

```{r}
food %>%
  mutate(sm_mean(calories))
```

As with any call to mutate, all of the original variables are kept in the output and
the new variable is added at the end. Using `select` we can verify that the average
calories has in fact been added to each row of the table.

```{r}
food %>%
  mutate(sm_mean(calories)) %>%
  select(item, food_group, calories, calories_mean)
```

The power of mutate summaries becomes particularly clear when grouping the
data. If we group the dataset by one or more variables and apply a summary
function within a mutation, the repeated summaries will be done within
each group. Here is an example of adding the average calories of each food
group to the dataset:

```{r}
food %>%
  group_by(food_group) %>%
  mutate(sm_mean(calories)) %>%
  select(item, food_group, calories, calories_mean)
```

Following this with a filter, for example, would allow us to select all of the foods
that have a less than average number of calories within their food group. We will see
many examples of grouped mutate summaries throughout the applications chapters.

## Exercises and References

There is are exercise notebook for this chapter:

- **exercies06A.Rmd**: practice modifying variables to a new dataset
- **exercies06B.Rmd**: open-ended questions to explore with data

It is strongly recommended to complete at least the first notebook prior to
continuing on to the next chapter.

The references mentioned at the end of chapter 5 contain useful information
about the mutate function itself. For more information about manipulating strings
and categorical variables with **stringi** and **forcats**, respectively, see:

- [R for Data Science: Chapter 14, Strings](https://r4ds.had.co.nz/strings.html)
- [R for Data Science: Chapter 15, Strings](https://r4ds.had.co.nz/factors.html)
- [stringi homepage](http://www.gagolewski.com/software/stringi/)
- [forcats cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/factors.pdf)

More details about string manipulation will be given in later chapters concerning
the processing of textual data and collecting data from the web.
