---
title: "09. Joining Data"
author: ""
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---


```{r, include=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)

food <- read_csv(file.path("data", "food.csv"))
food_prices <- read_csv(file.path("data", "food_prices.csv"))

food_sml <- food %>%
  select(item, food_group, calories)
```

## Relational data

A *primary key* consists of one or more variables that uniquely identify a row of
data. In our food dataset, the variable `item` is a primary key. Each row has a
unique value for this variable. Likewise, a *foreign key* is the appearance of a
primary key within a different dataset. The food group variable, for example, could
be a foreign key if we had another table with one row describing information about
each food group.

A primary key and the corresponding foreign key in another table form a *relation*.
Typically a relation maps a single row in one dataset to many rows in another.

A table join is a way of combining two tables based on relations. The goal is to
up a foreign key in one table with the primary key in another table, to add new variables
from one dataset into another dataset. Several examples will make this notion more clear.

## Joining data by key

As an example of performing table joins, we will start with a paired down version
of our foods data:

```{r, message=FALSE}
food_sml
```

Now, consider the following new dataset the corresponding dietary restrictions associated
with different food groups:

```{r, message=FALSE}
diet <- read_csv(file.path("data", "food_diet_restrictions.csv"))
diet
```

In the `diet` table, the food group variable is a primary key. Within the `food` table, it
is a foreign key. What we would like to do is to combine these datasets by
matching up rows that have the same values in the corresponding columns.

For example, it could be helpful to include the columns `vegan`, `vegitarian`, and `pescatarian`
in the foods dataset. In order to do this, we want to put the columns in the diet dataset into
the food dataset in a way that is similar to the `bind_cols` function. However, rather than
"lining up" the rows, we can to associate a row in the food dataset with the row in the diet
dataset that contains the same value of the variable `food_group`. The function to achieve
this is `left_join`; we pipe in the larger dataset that we want to join columns to, provide
the name of the table to grab columns from, and indicate what column will be used to combine
the two datasets.

```{r}
food_sml %>%
  left_join(diet, by = c("food_group"))
```

And that is it! We now have the dietary restrictions added into the food dataset. We can now
compule models, summaries, and construct visualizations based on the new metadata associated
with each food group.

Now, let's consider a third dataset containing information about food recipes, one for a Pot
Roast and another for Guacamole:

```{r, message=FALSE}
recipes <- read_csv(file.path("data", "food_recipes.csv"))
recipes
```

In this dataset, ingredient is a foreign key corresponding to the primary key "item"
in the foods dataset. For the recipes data, the primary key requires two columns:
both recipe and ingredient are needed to describe each row uniquely.

We can now try adding the calories information from the `food` dataset into the recipes dataset.
The complication here is that the column we want to join on has a different name in each dataset
("ingredient" versus "item"). To specify this, specify the two different names within the
`by =` argument of the join function:

```{r}
recipes %>%
  left_join(food_sml, by = c("ingredient" = "item"))
```

Now, we have a dataset that has added the food group and calories information to our recipe
dataset. Notice that the ingredient "Bay Leaf" is not in our food dataset. Because of this,
the food group and calories information for this ingredient are filled in with missing values.
This is similar to the behavior that we observed with `bind_rows` with variables that were
missing between the combined datasets.

Notice that rows of data about food items in the dataset `food` that  are not in any recipe
are not included in the output. This is where the term "left" comes from: all rows from the
"left" (e.g., first) table are returned, but non-matching rows in the second (e.g., "right")
table are dropped.

## Inner and outer joins

There are four mutating joins available in **dplyr**. These have the
same syntax. If the join key is a primary key in one table and there
are no missing matches, they all function exactly the same. If the
join key is a primary key in one table and not all foreign keys in
the other table match an entry in the other, the following happens:

- `left_join` only non-matching rows in the first dataset are kept
- `right_join` only non-matching rows in the second dataset are kept
- `full_join` non-matching rows are included from either table
- `inner_join` only matching rows from both tables are included

For example, by replacing `left_join` with the function `inner_join`
we can modify the join from the last section to drop the missing term
(Bay Leaf) that is not contained in our food dataset:

```{r}
recipes %>%
  inner_join(food_sml, by = c("ingredient" = "item"))
```

Here, now, the "Bay Leaf" row has been removed. Several other variations on join types
controlling for various kinds of dropping and keeping partial matches are provided by
the **dplyr** package. Details are linked to at the end of the chapter.

All of this can become quite complicated. I follow one simple rule
that make this much easier:

- make sure that the key used for the join is a primary key for
the *second* dataset; that is, it unique defines each row of the
second dataset (usually this dataset contains metadata about the
first dataset)

If you follow this rule, you only ever need `left_join` (if you are
okay with missing values in the metadata) and `inner_join`
(if you want to remove rows that did not have associated metadata).

## Exercises and References

There is are exercise notebook for this chapter:

- **exercies09A.Rmd**: practice applying table joins to a new dataset
- **exercies09B.Rmd**: practice and notes for applying table joins to a more complex dataset of airline codes

It is strongly recommended to complete at least the first notebook prior to
continuing on to the next chapter.

Here are several good resources if you want to learn more about relational data
in R:

- [R for Data Science: Chapter 13, Relational Data](https://r4ds.had.co.nz/relational-data.html)
- [dplyr cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)
- [dplyr homepage](https://dplyr.tidyverse.org/)

If you search online for help with R graphics, make sure to add the term "dplyr"
to your query as otherwise you may find advice for using other functions that
are not easily compatible with the approach described here.
