---
title: "20. Networks"
author: ""
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---


```{r, include=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(cleanNLP)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)

food <- read_csv(file.path("data", "food.csv"))
food_prices <- read_csv(file.path("data", "food_prices.csv"))
page_text <- read_csv(file.path("data", "page_text.csv"))
```

## Page Links

In this chapter we will start by working with network data taken from Wikipedia.
Specifically, we took all of the English Wikipedia pages associated with
each of the food items in the `foods` dataset. Wikipedia pages contain many
internal links between pages; we collected information about each time one of
the pages in our dataset provided a link to another page in our collection.
We will explore how to do produce this dataset using R code in Chapter 18.
Only links within the main body of the text were used.

The resulting dataset contains one row for each detected link. The two variables
in the dataset indicate the page the link was found on (`page`) and the name
of the page that was linked to. Here is the full dataset:

```{r, message=FALSE}
page_citation <- read_csv(file.path("data", "page_citation.csv"))
page_citation
```

Looking at the top of the dataset, which is arranged by the page name, we see that
the Wikipedia page for "Apple" has a link to "Banana" but no other pages. Asparagus,
on the other hand, links to five other food items.

We can use grouping and summarizing functions to see how often each page is linked
to. Arranging these in order gives some understanding of which pages are more
centrally located in our dataset.

```{r}
page_citation %>%
  group_by(link) %>%
  summarize(sm_count()) %>%
  arrange(desc(count))
```

However, counts can only get us so far. These say nothing, for example, about how easily
we can get hop between two pages by following 2, 3, or 4 links. In order to understand
the dataset we need a way of visualizing and modelling all of the connections at once.
This requires treating the dataset as a network (or graph).

A graph consists of a collection of objects (here, pages) and a collection of edges
between those objects (here, internal links).  An easy way to describe a network is
with an object that looks exactly like our dataset `page_citation`, called an edge
list, that describes pairs of objects that are linked together.

The **smodels** package provides an easy way of taking an edge list and returning
structured data the allows us to understand the dataset as a network. If we apply
the function `sm_graph_layout` to the edge list, an R object containing two elements
are returned. Splitting these off requires using the dollar sign notation, as follows,
and results in two datasets.

```{r}
graph <- sm_graph_layout(page_citation)
node <- graph$node
edge <- graph$edge
```

The node dataset contains extracted information about each of the objects in our
collection. We will describe each of these, in turn, throughout the remainder of
this chapter.

```{r}
node
```

The first column gives a label for the row. In the next two columns, named `x` and `y`,
is a computed way to layout the objects in two-dimensions that maximizes linked pages
being close to one another while minimizing the amount that all of the nodes are
bunched together. There is no exact meaning of the individual variables; rather, its
the relationships that they show that are interesting.

Using the first three variables, we can plot the pages as a scatter plot with labels
to see what pages appear to be closely related to one another.

```{r}
node %>%
  ggplot() +
    geom_point(aes(x = x, y = y)) +
    geom_text_repel(aes(x = x, y = y, label = id)) +
    theme_void()
```

Taking a look at the plot reveals some patterns that would match our intution. For example,
the seafood is mostly clustered together the upper-right of the plot and the vegatables are
together on the left of the plot. The relationships in this plot are learned by taking into
account which pages are linked to each other. It would be helpful to additionally put these
links onto the plot as well. This is where the `edge` dataset becomes useful.

The edge dataset contains one row for each edge in the dataset. The dataset has four columns
These describe the `x` and `y` values of one node in the edge and variables `xend` and `yend`
to indicate where in the scatterplot the ending point of the edge is.

```{r}
edge
```

We can add edges into the plot by adding a geom layer of type `geom_segment`. It takes
four aesthetics, named exactly the same as in `edge` dataset. The plot gets busy with
all of these lines, so we will set the opacity of them lower so as to not cluter the
visual space with the connections.

```{r}
node %>%
  ggplot(aes(x, y)) +
    geom_point() +
    geom_segment(aes(xend = xend, yend = yend), data = edge, alpha = 0.1) +
    geom_text_repel(aes(label = id)) +
    theme_void()
```

We now see the underlying data that describes the plot. Notice that here we are not
taking the direction of the links into account. In one of the subsequent sections
we will see how to plot the segements as arrows to indicate the direct that the links
occur in.

## Network Centrality

One of the core questions that arise when working with network data is trying to
identify the relative centrality of each node in the network. Several of the derived
measurements in the `nodes` dataset capture various forms of centrality. In this
section we will color the network diagram with these measurements and discuss what
properties each of them measures.

A component of a network is a collection of all the nodes that can be reached by
following along the edges. The `nodes` dataset contains an integer id called
`component` describing each of the components in the graph. These are ordered by
reverse size, so component 1 will always be the largest (or at least, tied for
the largest) component of the graph. Coloring the graph by component reveals
four part of the graph; we convert the component variable to a factor to force
ggplot to use a categorical scale.

```{r}
node %>%
  mutate(component = factor(component)) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend = xend, yend = yend), data = edge, alpha = 0.1) +
    geom_point(aes(color = component), size = 5) +
    theme_void()
```

This component number gives a very coarse measurement of how central a node is. We
can say that, in some way, all of the nodes in component 1 are more central than
the other 6 orphaned pages. Note that having one large component and several small ones
is a typical features of many naturally occuring networks.

Another measurement of centrality is a node's degree. The degree of a node
is the number of neighbors it has; in other words, it counts how many edges
the node is a part of. The degree of each node has been computed in the `node`
table. This is similar to the counts that be produced in the
first section by counting occurances in the raw edge list. Ploting degree on
our plot shows that the nodes with the highest degree do seem to sit in the
middle of the plot and correspond to a high number of having a large number
of edges.

```{r}
node %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend = xend, yend = yend), data = edge, alpha = 0.1) +
    geom_point(aes(color = degree), size = 5) +
    scale_color_binned(type = "viridis", n.breaks = 15) +
    theme_void()
```

As mentioned above, the degree of a node does not take into account anything
other than direct neighbors. A more holistic measurement is given by a quantity
called the *eigenvalue centrality*, which is provided in the `node` table. This
centrality value provides a score for each node that is proportional to the sum
of the scores of its neighbors. Mathematically, it assigns a set of scores $s_j$
for each node such that:

$$ s_j = \lambda \cdot \sum_{i \in \text{Neighbors{j}}} s_i $$

If you have taken a course in linear algebra and are familiar with the eigenvalues
of a matrix, know that the score can be efficently calculated by an eigenvalue
equation. If not, just know that it's a score that takes into account all of the
nodes at once. The eigenvalue score is, by convention, scales so that the largest
score is 1. It is only possible to describe the eigenvalue centrality scores for
a connected set of nodes on a graph, so the computation is done individually for
each component. Plotting the scores on the plot shows a slightly different pattern
compared to the degree centrality scores.

```{r}
node %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend = xend, yend = yend), data = edge, alpha = 0.1) +
    geom_point(aes(color = eigen), size = 5) +
    scale_color_binned(type = "viridis", n.breaks = 15) +
    theme_void()
```

The biggest difference is that the eigenvalue centrality is more concentrated
on one set of connections, whereas degree centrality is more spread out.

Another measurement of centrality is given by the *closeness centrality* score.
For each node in the graph, consider the minimim number of edges that are needed
to from this node to any other node within its component. Adding the recipricol
of these scores together gives a measurment of how close a node is to all of the
other nodes in the graph. This is the *closeness centrality* score for a node
and is given as the variable `close` in our table. Notice that the scale of the
closeness centrality score will be very different depending on the size of a
component, so in our plot we will filter to include only those nodes in the
largest component.

```{r}
node %>%
  filter(component %in% c(1)) %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend = xend, yend = yend), data = edge, alpha = 0.1) +
    geom_point(aes(color = close), size = 5) +
    scale_color_viridis_c() +
    theme_void()
```

Notice that these scores are slightly different than the eigenvalue centrality
scores, with the most central node moving from the "Lemon" page to the "Tomato"
page.

The final measurement of centrality we have in our table, *betweenness centrality*
also comes from considering minimal paths. For every two nodes in a connected
component, consider all of the possible ways to go from one to the other along
edges in the graph. Then, consider all of the paths (there may be only one)
between the two nodes that require a minimal number of hops. The betweenness
centrality scores measures how many of these minimal paths go through each node
(there is some normalization to account for the case when there are many minimal
paths, so the counts are not exact integers). This score is stored in the variable
`between`.

```{r}
node %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend = xend, yend = yend), data = edge, alpha = 0.1) +
    geom_point(aes(color = between), size = 5) +
    scale_color_viridis_c() +
    theme_void()
```

The betweeness score here, as is often the case, tends to be have a different
pattern than the other centrality scores. It gives a high score to bridges
between different parts of the network, rather than giving high weight to
how central a node is within a particular cluster. In the next section we will
see that this is a useful property for understanding the structure of a network.

## Centrality Applications

While it is instructive to color the graph plots above with the centrality scores,
most applications of graph centrality focus on studying the `node` table directly
to understand the structure created by a set of relationships. For example, we
could sort the node table by component and eigenvalue scores to get a sense of
what pages are the most central to the graph. Here, for example, we see that
the eigenvalue centrality is largest for the citrus fruits ("Lemon", "Lime",
"Grapefruit", and "Orange"):

```{r}
node %>%
  arrange(component, desc(eigen))
```

Looking at a two-way relationship, we can dig deeper into the way that the graph is structured.
For example, take a plot of the eigenvalue and betweeness scores for the pages:

```{r}
node %>%
  filter(component %in% c(1)) %>%
  ggplot() +
    geom_point(aes(x = eigen, y = between)) +
    geom_text_repel(aes(x = eigen, y = between, label = id))
```

The pages with a high betweeness score but relatively low eigenvalue score ("Asparagus",
"Shrimp", "Potato", "Tomato", "Lobster") indicate pages that links different parts of the
graph together. Tomato is often thought of as sitting inbetween fruits and vegetables,
for example. The Asparagus happens to mention that it is frequently "stir-fried with
chicken, shrimp, or beef", causing it to be a gatekeeper between vegetables and the
meats.

Plotting eigenvalue centrality and closeness reveals additional patterns in the dataset.

```{r}
node %>%
  filter(component %in% c(1)) %>%
  ggplot() +
    geom_point(aes(x = eigen, y = close)) +
    geom_text_repel(aes(x = eigen, y = close, label = id))
```

As seen in the layout plots, this graph shows that the citrus fruits have very high eigenvalue
scores but only modest closeness. Nodes that have high closeness scores but modest eigenvalue
scores include "Beef" and "Yogurt", which are central to their only clusters of the meats and
diary products, respectively.

## Network Clusters

Another algorithm that we can perform on a graph is that of clustering. Here,
we try to split the nodes into groups such that a large number of the edges are
between nodes within a group rather than across groups. Here is a visualization
of the clusters detected on our graph:

```{r}
node %>%
  ggplot(aes(x, y)) +
    geom_segment(aes(xend = xend, yend = yend), data = edge, alpha = 0.1) +
    geom_point(aes(color = cluster), size = 5) +
    theme_void()
```

We can paste together the nodes within a cluster to try to further understand the
internal structure of the relationships.

```{r}
node %>%
  filter(component %in% 1) %>%
  group_by(cluster) %>%
  summarize(sm_paste(id)) %>%
  select(id_paste)
```

Taking our knowledge of the underlying food types, these clusters line up well with
our understanding of what food types would be related to one another. The fish are
grouped into two clusters. The shelfish are placed in a different cluster (with only
Lobster left-out and placed with the fish). All of the citric fruits are put together.
The third cluster is a bit more esoteric, as it contains some vegetables, some
meat, and some dairy products.

## Directed Networks

At the start of the chapter, we noted that our graph assumed that edges did not have
a direction. It is possible to compute graph metrics that take this relationship into
account by setting the `directed` argument of `sm_graph_layout` to `TRUE`.

```{r}
graph <- sm_graph_layout(page_citation, directed = TRUE)
node <- graph$node
edge <- graph$edge
```

The `node` table contains a slightly different set of measurements. Closeness
centrality is no longer available; eigenvalue centrality is still computed,
but is done so without using the directions of the edges. There are now three
different degree counts: the out-degree (number of links on the page) the
in-degree (number of links into a page), and the total of these two.

```{r}
node
```

To visualize the directed graph, we add an `arrow` argument to the `geom_segement`
layer. This shows, for example, that "Peach" and "Plum" both link to each other
whereas "Salmon" links to "Catfish", but not vice-versa.

```{r}
node %>%
  ggplot(aes(x, y)) +
    geom_point(alpha = 0.2) +
    geom_text_repel(aes(label = id), alpha = 0.5) +
    geom_segment(
      aes(xend = xend, yend = yend),
      data = edge,
      alpha = 0.7,
      arrow = arrow(length = unit(0.02, "npc"))
    ) +
    theme_void()
```

An interesting plot with the data from the directed graph is to show the in-degree
as a function of the out-degree.

```{r}
node %>%
  filter(component %in% c(1)) %>%
  ggplot() +
    geom_point(aes(x = degree_out, y = degree_in)) +
    geom_text_repel(aes(x = degree_out, y = degree_in, label = id))
```

This shows that some pages from our last analysis, such as "Tomato", are primarily
central because they link to many other pages. In comparison, other central pages
such as "Lime" and "Apple" are mostly connected by links into the pages.

## Cocitation Network

The graph structure we have been working with is a form of graph called a "citation
network". Pages are join whenever one links to another. This is a popular method in
understanding academic articles, friendship networks on social media (i.e., tracking
mentions on Twitter), or investigating the relative importance of court cases. There
are some drawbacks of using citation counts, however. They are sensistive to the
time-order of publication, they are effected by the relative length of each document,
and they are easily effected by small changes. For example, we above saw the large
effect that one sentence about stir-fry dishes on the "Asparagus" page had in
increasing that page's overall centrality score.

A co-citation network is a method of showing links across a citation network while
avoiding some of the pitfalls that arise when using direct links. A co-citation is
formed between two pages whenever a third entry cites *both* of them. The idea is
that if a third source talks about two sources in the same reference, there is
likely a relationship between the documents.

We created a co-citation dataset from Wikipedia by first downloading all of the
pages linked to from any food pages in our collection. We then counted how often
any pair of pages in our dataset were both linked to from the same source. The
dataset is available here; duplicate rows were not removed, but the `page1` and
`page2` were sorted so that `page1` always comes alphabetically before `page2`.

```{r, message=FALSE}
page_cocitation <- read_csv(file.path("data", "page_cocitation.csv"))
page_cocitation
```

Counting the number of times each pair occurs can be done by grouping the table
and using the `sm_count` summary function.

```{r}
page_cocitation %>%
  group_by(page1, page2) %>%
  summarize(sm_count())
```

When working with co-citations, it is useful to only include links between two
pages when the count is above some threshold. Here, we use a cut-off of 6 pages.
Notice that co-citations are fundamentally undirected relationships.

```{r}
graph <- page_cocitation %>%
  group_by(page1, page2) %>%
  summarize(sm_count()) %>%
  filter(count >= 6) %>%
  sm_graph_layout()

node <- graph$node
edge <- graph$edge
```

Plotting the graph, we see that there is a very different structure compared to
the citation plot.

```{r}
node %>%
  ggplot(aes(x, y)) +
    geom_point() +
    geom_segment(aes(xend = xend, yend = yend), data = edge, alpha = 0.1) +
    geom_text_repel(aes(label = id)) +
    theme_void()
```

In the largest component, there are more links and less clustering
across types of foods. The exercies include more examples to try to
become familiar with the structure of these co-citation networks and
how they tend to differ from citation networks.

## Distance Network

```{r, message=FALSE}
token <- read_rds("cache/tokens.rds")
edges <- token %>%
  cnlp_utils_tfidf(token_var = "lemma") %>%
  sm_tidy_angle_distance()  %>%
  filter(document1 != document2) %>%
  group_by(document1) %>%
  arrange(distance) %>%
  slice_head(n = 3) %>%
  select(document1, document2)
```

A final type of network that we will explore here is a distance network. We
start by computing the angle distance between the term-frequency scores for
the text contained on each food's Wikipedia page (see chapter 11). We then
select edges by taking the closest 100 pairs of pages based on the texts.

```{r}
edges <- token %>%
  cnlp_utils_tfidf(token_var = "lemma") %>%
  sm_tidy_angle_distance()  %>%
  filter(document1 != document2) %>%
  arrange(distance) %>%
  slice_head(n = 100) %>%
  select(document1, document2)
```

Using these edges, we can create a graph network. As with co-citations, it would
not make sense to use a directed network for the distance-based graph.

```{r}
graph <- sm_graph_layout(edges)
node <- graph$node
edge <- graph$edge
```

Looking at the clusters formed by the distance graph shows a grouping of
the foods that very closely matches our understanding of the food types:

```{r}
node %>%
  group_by(cluster) %>%
  summarize(sm_paste(id))
```

Plotting the network shows that the structure is, again, different than the
citation and co-citation networks. Rather than one large dominant component,
there are several smaller, but tightly connected, components.

```{r}
node %>%
  ggplot(aes(x, y)) +
    geom_point() +
    geom_segment(aes(xend = xend, yend = yend), data = edge, alpha = 0.1) +
    geom_text_repel(aes(label = id)) +
    theme_void()
```

After axploring many examples of citation, co-citation, and distance networks,
the typically patterns of each will become clear. Often the structure of a large
network becomes apparent only after looking at several different versions of
possible graphs.
