---
title: "Notebook 11"
output: html_document
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
```

## Movies Data

Over the next few classes we will be working with a dataset of movies I have
constructed consisting of the top 100 grossing films for each year from 1970 to
2019. The data comes from IMDb. Today we will focus on getting familiar with
the various components of the data. Let's read in the four tables of data,
as well as a data dictionary, and then go through each of the tables.

```{r, message=FALSE}
movies <- read_csv(file.path("data", "movies_50_years.csv"))
m_genre <- read_csv(file.path("data", "movies_50_years_genre.csv"))
m_people <- read_csv(file.path("data", "movies_50_years_people.csv"))
m_dict <- read_csv(file.path("data", "movies_50_years_data_dictionary.csv"))
m_color <- read_csv(file.path("data", "movies_50_years_color.csv"))
```

See notebook11 and the data dictionary for more information about the available
variables.

### Movie Color

Start by verifying that the percentage values for each film each add up to 100
(there may be some slight rounding error, but everything should be very close
to 100). Use whatever method you find to be the easiest or most reliable, but
do not resort to manually checking the values for each film.

```{r}
m_color %>%
  group_by(year, title) %>%
  summarize(sm_sum(percentage)) %>%
  ungroup() %>%
  summarize(sm_min(percentage_sum), sm_max(percentage_sum))
```

Next, associate each film to the color that is most dominent in each film
poster. Plot a count of the most dominant colors using a bar plot. No need for
any labels, titles, or other finishing touches, but do try to color the bars
according to the associated color name. Note that you will have to use the
aesthetic "fill" and scale `scale_fill_identity` to do this.

```{r}
m_color %>%
  group_by(year, title) %>%
  arrange(desc(percentage)) %>%
  slice(1) %>%
  group_by(color) %>%
  summarize(sm_count()) %>%
  ggplot(aes(color, count)) +
    geom_col(aes(fill = color), color = "black") +
    scale_fill_identity()
```

Redo the same below, but now include the **two** most dominant colors. Note any
patterns that you see in relation to the first plot. Note: this should be any
easy tweak of your last plot.

```{r, question-01}

```

Finally, filter out just those rows where `color_type` is hue and use
a wide pivot to create a dataset with one row per film and a column
associated with each color. Pipe this into a plot to show the relationship
between the red intensity and the orange intensity.

```{r, question-02}

```

### Movie People

Compute the number of people listed as starring in each film, compute the
average number of people listed as starring in a film for each year, and
plot the pattern over the 50 years of data that we have available to us.
Do you notice anything strange about the dataset?

```{r, question-03}

```

Filter the data to those names where the gender confidence score is
less than 0.6. Note any patterns that you see and consider caveats that
any gender-based analysis on the larger dataset should consider.

```{r, question-04}

```

Now, make a plot showing the number of films starring by the 20 most prolific
directors with the bars filled according to an actor's gender. Note, consider
grouping the data by both gender and person before doing the summarization.

```{r, question-05}

```

You will (hopefully) notice something strange in the plot above. Fix this by
only including actors with a high gender confidence score (above 0.95 perhaps?).

```{r, question-06}

```
