---
title: "Notebook 16"
output: html_document
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(lubridate)
library(sf)
library(units)
library(RcppRoll)
library(hms)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
options(lubridate.week.start = 1)
Sys.setlocale(locale = "en_US.UTF-8")

sm_centroid <- function(data) {
  suppressWarnings({ z <- st_coordinates(st_centroid(data)) })
  return(tibble(lon = z[,1], lat = z[,2]))
}
spatial_join <- function(...) {
  return(st_as_sf(as_tibble(st_join(...))))
}
```

## Chicago Data

### Load the Data

Let's load the data that we will be looking at through the remainder of the
semester:

```{r, message = FALSE}
comarea <- read_sf(file.path("data", "chicago_community_areas.geojson"))
ziparea <- read_sf(file.path("data", "zip_codes.geojson"))
socio <- read_csv(file.path("data", "census_socioeconomic.csv"))
medical <- read_csv(file.path("data", "chicago_medical_examiner_cases.csv.gz"))
```

This time, we will look into the temporal components of the data
and see how they can be integrated into the spatial visualisations.

## Time and Datetime objects

### What is a datetime object?

The medical examiner data has two fields that describe specific times,
the time of the incident and the time of death. We call these *datetime*
objects because they describe a specific time on a specific day.

```{r}
medical
```

Notice that these are similar but a bit different than the date objects we had
in the previous notes. As with dates, there exist several functions to extract
particular components of datetime objects. These include all of the functions
for dates, as well as specific ones for time of day:

```{r}
medical %>%
  select(date_incident_iso) %>%
  mutate(
    year = year(date_incident_iso),
    hour = hour(date_incident_iso),
    minute = minute(date_incident_iso)
  )
```

Sometimes datetime data is given overly precisely, making it hard to
group and count the data. A helpful function to working with such
data is `floor_date`, which rounds down to the nearest interval:

```{r}
medical %>%
  select(date_death_iso) %>%
  mutate(hour_death_iso = floor_date(date_death_iso, "hour"))
```

You can use different intervals---such as "minute", "day", "week",
or "2 hour"---depending on your specific application.

### Creating datetime objects

As with date objects, it can be useful to create datetime values on
the fly. This can be done by specifying individual components with
`make_datetime` (defaults go to 1970-01-01 00:00):

```{r}
medical %>%
  arrange(date_incident_iso) %>%
  filter(date_incident_iso > make_datetime(2020, 3, 1, 5, 6))
```

Or specifying the entire value as a string:

```{r}
medical %>%
  arrange(date_incident_iso) %>%
  filter(date_incident_iso > ymd_hms("2020-03-01 05:06:00"))
```

These can be useful for manipulating existing dates or for filtering data.

### Time Differences

Adding (or subtracting) integers from a date object increases (or decreases)
the date by the given number of days. A datetime object works similarly, but
changes by a number of seconds:

```{r}
medical %>%
  select(date_incident_iso) %>%
  mutate(date_incident_iso_plus1 = date_incident_iso + 1)
```

We can also take the difference between two datetime objects; converting
the result to a number with `as.numeric` yields the number of seconds
between the time time periods. Here is the difference between the incident
datetime and the death datetime:

```{r}
medical %>%
  select(date_incident_iso, date_death_iso) %>%
  mutate(diff = as.numeric(date_death_iso - date_incident_iso))
```

### Time Objects

Finally, we also have the ability to create a *time* object. This records
a time without a corresponding date. It can be useful for when you want
arithmetic to behave properly (that is, wrap around at midnight) or if
you want to group/plot datetime objects using only their time component.

The function `as_hms` will create a time object from a date object:

```{r}
medical %>%
  select(date_death_iso) %>%
  mutate(time_death = as_hms(date_death_iso))
```

Finally, the function hms can be used to create a *time* object from
scratch.

### Application

Let's see how to put some of these elements together. In the plot below
we show the number opioiod related and non-opioid related deaths investigated
by the medical examiner's office by the hour of the day:

```{r}
medical %>%
  mutate(hour_death_iso = floor_date(date_death_iso, "hour")) %>%
  mutate(hour_death_iso = as_hms(hour_death_iso)) %>%
  group_by(hour_death_iso, opioid_related) %>%
  summarize(sm_count()) %>%
  ggplot(aes(hour_death_iso, count)) +
    geom_point(aes(color = opioid_related)) +
    geom_line(aes(color = opioid_related)) +
    scale_y_continuous(limits = c(0, NA))
```

## Practice

### Time of Day: Opioid Related

Repeat the code in the application above, but show the number of incidents
(rather than deaths) that occur at each hour of the day, by whether the incident
was opioid related or not.

```{r, question-01}

```

Do you notice anything strange? What percentage of non-opioid related deaths have
an incident recorded exactly at midnight?

```{r, question-02}

```

Does there seem to be a data issue here?

### Time to Death

Compute the time between the recorded death and the recorded incident. Order
the data with the longest duration at the top. Do you see anything strange?
Is there a data issue here?

```{r, question-03}

```

I want you to investiage the median duration between the time of death and the
time of the incident within the top-100 most common `primary_cause` labels.
Create a table with the median time for each of the top-100 categories and
arrange the table from the shortest to the longest duration

```{r, question-04}

```

Note: If you have the difference in time in seconds, try using the function `as_hms`
to convert these integers into hours, minutes, and seconds. Do the results seem
reasonable to you? Any strange outliers?

### Heat and Cold Deaths

Create a plot with hour on the x-axis and month on the y-axis. Show a scatter plot
with points showing the total number of cold-related deaths that occured (incident time,
not time of death) at each combination of hour of the day and month. Try to make the
plot look nice (no need for fancy labels, but maybe make sure to include real month
names and properly formatted dates).

```{r, question-05}

```

Repeat the previous question, but look at heat related deaths.

```{r, question-06}

```

Do the patterns surprise you at all?

### Time of the Week

In this final question, I want you to show the number of deaths on the y-axis as a
function of the hour of the week (hour 0 is Monday at midnight, hour 24 is Tuesday
at midnight, etc). It will take a little bit of work to get this correct, particularly
if you want to axes to be labelled correctly without resorting to hand labels.

```{r, question-07}

```
