---
title: "Notebook 09"
output: html_document
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)

food <- read_csv(file.path("data", "food.csv"))
food_prices <- read_csv(file.path("data", "food_prices.csv"))
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

# Relational data

A *primary key* consists of one or more variables that uniquely identify a row
of data. In our food dataset, the variable `item` is a primary key. Each row
has a unique value for this variable. Likewise, a *foreign key* is the
appearance of a primary key within a different dataset. The food group variable,
for example, could be a foreign key if we had another table with one row
describing information about each food group.

A primary key and the corresponding foreign key in another table form a
*relation*. Typically a relation maps a single row in one dataset to many rows
in another.

A table join is a way of combining two tables based on relations. The goal is
to match up a foreign key in one table with the primary key in another table,
to add new variables from one dataset into another dataset. Several examples
will make this notion more clear.

## Joining data by key

As an example of performing table joins, we will start with a paired down
version of our foods data:

```{r, message=FALSE}
food_sml <- food %>%
  select(item, food_group, calories)

food_sml
```

Now, consider the following new dataset corresponding to dietary restrictions
associated with different food groups:

```{r, message=FALSE}
diet <- read_csv(file.path("data", "food_diet_restrictions.csv"))
diet
```

In the `diet` table, the food group variable is a primary key. Within the
`food` table, it is a foreign key. What we would like to do is to combine these
datasets by matching up rows that have the same values in the corresponding
columns.

For example, it could be helpful to include the columns `vegan`, `vegetarian`,
and `pescatarian` in the foods dataset. In order to do this, we need to
associate a row in the food dataset with the row in the diet dataset that
contains the same value of the variable `food_group`. We do this using the
function `left_join`; we pipe in the larger dataset that we want to join columns
to, provide the name of the table to grab columns from, and indicate what
column will be used to combine the two datasets.

```{r}
food_sml %>%
  left_join(diet, by = c("food_group"))
```

And that is it! We now have the dietary restrictions added into the food
dataset. We can now compute models, summaries, and construct visualizations
based on the new metadata associated with each food group.

Now, let's consider a third dataset containing information about food recipes,
one for a Pot Roast and another for Guacamole:

```{r, message=FALSE}
recipes <- read_csv(file.path("data", "food_recipes.csv"))
recipes
```

In this dataset, ingredient is a foreign key corresponding to the primary key
"item" in the foods dataset. For the recipes data, the primary key requires two
columns: both recipe and ingredient are needed to describe each row uniquely.

We can now try adding the calories information from the `food` dataset into the
recipes dataset. The complication here is that the column we want to join on has
a different name in each dataset ("ingredient" versus "item"). To specify this,
specify the two different names within the `by =` argument of the join function:

```{r}
recipes %>%
  left_join(food_sml, by = c("ingredient" = "item"))
```

Now, we have a dataset that has added the food group and calories information
to our recipe dataset. Notice that the ingredient "Bay Leaf" is not in our food
dataset. Because of this, the food group and calories information for this
ingredient are filled in with missing values.

Notice that rows of data about food items in the dataset `food` that  are not
in any recipe are not included in the output. This is where the term "left"
comes from: all rows from the "left" (e.g., first) table are returned, but
non-matching rows in the second (e.g., "right") table are dropped.

## Inner and outer joins

There are four mutating joins available in **dplyr**. These have the
same syntax. If the join key is a primary key in one table and there
are no missing matches, they all function exactly the same. If the
join key is a primary key in one table and not all foreign keys in
the other table match an entry in the other, the following happens:

- `left_join` only non-matching rows in the first dataset are kept
- `right_join` only non-matching rows in the second dataset are kept
- `full_join` non-matching rows are included from either table
- `inner_join` only matching rows from both tables are included

For example, by replacing `left_join` with the function `inner_join`
we can modify the join from the last section to drop the missing term
(Bay Leaf) that is not contained in our food dataset:

```{r}
recipes %>%
  inner_join(food_sml, by = c("ingredient" = "item"))
```

These four joins are called "mutating" joins because, like the mutate function,
they add columns to a dataset. There are also two "filter" joins that serve only
to select rows in the first (left) table. The function `semi_join` selects
rows in the first table that has a match in the second and `anti_join` selects
rows in the first table that do not have a match in the second. These are
useful for several data science tasks and will come up from time-to-time in our
applications.

Finally, there is another join called a "nested" join (`nest_join`). We may use
this when doing web scraping towards the end of the semester, but we likely
will avoid it until then as it is a bit more complicated.

# Practice

## Largest City and Köppen Datasets

In this notebook we will work with the largest cities dataset:

```{r, message=FALSE}
cities <- read_csv(file.path("data", "largest_cities.csv"))
```

We will also load a dataset containing metadata about the Köppen climate
classification scores.

```{r, message=FALSE}
koppen <- read_csv(file.path("data", "koppen_metadata.csv"))
koppen
```

## Joining Climate Data

In the code block below, I am using the `left_join` function to combine the
`cities` dataset with the `koppen` metadata. Note that the key has a different
name in each of the two datasets.

```{r}
cities %>%
  left_join(koppen, by = c("koppen_code" = "code"))
```

Take a close look at the dataset you created in the last question. Scrolling
all the way to the final columns, you should see the name of the climate region
and the default color added into the dataset.

Something, however, has happened to the first variable in the combined
dataset. What is the (new) name of the first variable in the combined
output? **Answer**:

It turns out that both the `cities` dataset and the `koppen` dataset both
contain a variable named "name". In R, a dataset cannot contain two columns
with the same names. By default, `left_join` adds a suffix to distinguish
variables that came from the first dataset and those that came from the
second dataset. I generally find the default suffix values to not be
ideal because they change both the names in the left table (the one that
contains the observations of interest) as well as the the new metadata
columns. Let's see how to change these suffix values.

The `left_join` function has an optional argument called `suffix`. You
can supply two string values to it to describe what suffix values are
placed on the first and second tables in the join. The default value is
`c(".x", ".y")`. I typically find something with the first entry empty
to be more practical. So, for example: `c("", "_koppen")`. In the code
below, repeat the left join with this suffix option:

```{r, question-01}

```

Verify that the variable names match what you expected to find in the new
dataset. The same suffix option can be used to modify the output for other
table joins.

Below, pipe the joined dataset from the last question into a plot with
longitude on the x-axis and latitude on the y-axis. Color the points based on
the default Köppen codes using an identity color scale:

```{r, question-02}

```

## Creating join table

So far we have worked only with data loaded into R through csv or Excel files.
It is also possible to directly create datasets within R code. This can be very
useful for quickly creating small metadata tables, particularly as a way of
cleaning and labelling data. In order to make a new dataset directly in R we
use the `tibble` function, describing each variable with names followed by
equals signs and the values contained within the function `c()`. For example,
here is a dataset with a column giving the names of the five continents in
our dataset next to the names of the continents in Georgian:

```{r}
continent_names <- tibble(
  continent = c("Asia", "South America", "North America", "Africa", "Europe"),
  continent_ka = c("აზია", "სამხრეთი ამერიკა", "ჩრდილოეთი ამერიკა", "აფრიკა", "ევროპა")
)

continent_names
```

In the code below, join the `cities` dataset with the `continent_ka` dataset.
Pipe the data into a plot with longitude on the x-axis and latitude on the
y-axis, coloring the points with the variable `continent_ka`.

```{r, question-03}

```

The plot should now show the legend in Georgian. In the code below, repeat
the procedure with a different language of your choice and recreate the
plot with the new labels.

```{r, question-04}

```

Note that these plots may not work locally on your own machine, depending on
the character set you choose. They should, however, run on the Linux server
running RStudio Cloud.

## Manual colors with a legend

There is one problem with the plot that we created to color the
cities by their Köppen codes. There's no legend for the plot. A
way to rectify this problem is to color the points using the
variable `name_koppen` and use the colors in the `koppen` metadata
table to describe a manual scale with `scale_color_manual`.
Specifically, we need:

   `scale_color_manual(values = koppen$color, labels = koppen$name)`.

Below, redo your plot with this color value and scale.

```{r, question-05}

```

The plot should like nice, but does requires a wide screen to see all of the
colors given the long descriptions in the legend.
