---
title: "Notebook 15"
output: html_document
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(lubridate)
library(sf)
library(units)
library(RcppRoll)
library(hms)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
options(lubridate.week.start = 1)
Sys.setlocale(locale = "en_US.UTF-8")

sm_centroid <- function(data) {
  suppressWarnings({ z <- st_coordinates(st_centroid(data)) })
  return(tibble(lon = z[,1], lat = z[,2]))
}
spatial_join <- function(...) {
  return(st_as_sf(as_tibble(st_join(...))))
}
```

## Chicago Data

### Load the Data

Let's load the data that we will be looking at through the remainder of the
semester:

```{r, message = FALSE}
comarea <- read_sf(file.path("data", "chicago_community_areas.geojson"))
ziparea <- read_sf(file.path("data", "zip_codes.geojson"))
socio <- read_csv(file.path("data", "census_socioeconomic.csv"))
medical <- read_csv(file.path("data", "chicago_medical_examiner_cases.csv.gz"))
```

This time, we will look into the temporal components of the data
and see how they can be integrated into the spatial visualisations.

## Time and Datetime objects

### What is a date time object?

```{r}
medical
```

```{r}
medical %>%
  select(date_incident_iso) %>%
  mutate(year = year(date_incident_iso), hour = hour(date_incident_iso), minute = minute(date_incident_iso))
```

```{r}
medical %>%
  select(date_incident_iso) %>%
  mutate(date_incident_iso_plus1 = date_incident_iso + 1)
```

```{r}
medical %>%
  select(date_incident_iso, date_death_iso) %>%
  mutate(diff = as.numeric(date_death_iso - date_incident_iso))
```


```{r}
medical %>%
  select(date_death_iso) %>%
  mutate(time_death = as_hms(date_death_iso))
```

```{r}
medical %>%
  select(date_death_iso) %>%
  mutate(hour_death_iso = floor_date(date_death_iso, "hour")) 
```

```{r}
medical %>%
  mutate(hour_death_iso = floor_date(date_death_iso, "hour")) %>%
  mutate(hour_death_iso = as_hms(hour_death_iso)) %>%
  group_by(hour_death_iso, opioid_related) %>%
  summarize(sm_count()) %>%
  ggplot(aes(hour_death_iso, count)) +
    geom_point(aes(color = opioid_related)) +
    geom_line(aes(color = opioid_related)) +
    scale_y_continuous(limits = c(0, 2000))
```

## Practice

