---
title: "Notebook 13"
output: html_document
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(sf)
library(units)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)

sm_centroid <- function(data) {
  suppressWarnings({ z <- st_coordinates(st_centroid(data)) })
  return(tibble(lon = z[,1], lat = z[,2]))
}
```

We also need another package that was not originally installed. Uncomment
the following line, run it, and then re-comment it.

```{r}
#install.packages("lwgeom")
```

It does not need to be directly loaded, but is needed "under the hood" by some
of the other spatial packages.

## French COVID-19 Data

### Overview

For our next unit, we will be looking at a collection of spatio-temporal data
(that is, data with a time component and spatial component) concerning the
ongoing COVID-19 pandemic. We will start by looking at data from France, and
then move to data from the United States. Today we will introduce tools for
working with spatial data. In the next notebook we will see techniques for
working with data containing a date variable.

Here are the three French datasets that we will be working with. They contain
spatial information, population metadata, and coronavirus numbers at the level
of French *départements*. These are geographic areas that are an important
political entities in France. There are 101 départements; 96 in mainland Europe
and 5 overseas (called the *DOM*, or  Départements d'outre-mer).

```{r, message = FALSE}
dept <- read_sf(file.path("data", "france_departement.geojson"))
pop <- read_csv(file.path("data", "france_departement_population.csv"))
covid <- read_csv(file.path("data", "france_departement_covid.csv"))
```

The coronavirus data is stored with one row for each combination of day and
département. We have the cumulative number of people who died in each day from
COVID-19, the total number currently in hospital, the total number currently
in *reanimation* (this is similar to ICU, but not exactly equivalent, so I
used the french term here), and the cumulative number of newly recovered. Notice
that deceased and recovered are the *total* counts of people who have died or
recovered, whereas hospitalised and reanimation are the numbers at that moment
of patients in each group. There are columns indicating the number of new
hospitalisations and reanimations, but these have many missing data points.

```{r}
covid
```

Note that, along with date, either the `departement` or `departement_name` can
be used as a primary key for the data. You only need one to uniquely describe a
location.

Unlike the United States, France collects and publishes very little demographic
data about its citizens. One of the few variables we will be able to look at for
each département is its population, which is in the following table:

```{r}
pop
```

When working with the county-level U.S. data for project 3, you will have more
demographic variables to work with.

### Working with spatial data

We also have loaded spatial data about each département in France in the form of
a *simple feature collection*. The data was loaded from a "geojson" file: a
plain-text, open specification for describing spatial data and associated
metadata. Printing out the dataset shows that it is not too different from an
"ordinary" table of data:

```{r}
dept
```

Like the `pop` dataset, there is one row for each département. It has sole extra
metadata (printed in a different RStudio window) and a special column called
`geometry`. The geometry holds all of the information indicating *where* the
associate geographic area is on a map.

Most plotting, data manipulation, and modeling functions can be used with a
spatial data frame just the same way we used plain data frame. For example, we
can do a left join with the population data and slice off the first 96 rows
(these are the areas that are in Europe).

```{r}
dept %>%
  left_join(pop, by = "departement") %>%
  slice(1:96)
```

Notice that the spatial components of the data frame are still present after
joining and slicing the data.

Standard **ggplot** functions work to visualise the non-spatial components of
our spatial data. To show the spatial component we need to use a unique kind
of geometry called `geom_sf`. It will plot the shapes in the dataset (by default
from the `geometry` column) over a map. Here is an example of France using
our spatial data:

```{r}
dept %>%
  slice(1:96) %>%
  ggplot() +
    geom_sf()
```

We can control the way the map looks by adjusting the aesthetics, just as with
any other geometry:

  - color (border color)
  - fill (interior color)
  - size (width of the border)
  - alpha (transparency of the shapes)

Here, we will make the borders very small and show the overall population of
each département:

```{r}
dept %>%
  left_join(pop, by = "departement") %>%
  slice(1:96) %>%
  ggplot() +
    geom_sf(aes(fill = population), color = "black", size = 0.1) +
    scale_fill_viridis_c()
```

You may disagree, but I think that's a pretty nice map with not too much
extra work! While you may be surprised to see this, the largest population
is in fact in the Nord département and not Paris (though the latter has
a much higher population density).

Two other spatial geometries exist: `geom_sf_text` and `geom_sf_label` for
adding labels to a plot. For example, here we can name some of the areas:

```{r, warning=FALSE}
dept %>%
  left_join(pop, by = "departement") %>%
  slice(1:96) %>%
  ggplot() +
    geom_sf(color = "black", fill = "white", alpha = 0.4, size = 0.1) +
    geom_sf_text(aes(label = departement_name), check_overlap = TRUE, size = 2)
```

Note that some areas are not labelled because we set `check_overlap` to `TRUE`.

### Spatial operations

We can also use the spatial information in our dataset to compute metrics about
the geometric areas. For example, the function `st_area` computes the total area
of each value in the geometry column (two extra functions are needed to convert
the output to a usable number of square-kilometers).

```{r}
dept %>%
  mutate(area = as.numeric(set_units(st_area(geometry), "km^2")))
```

From there, we could join to the population data and compute the population
density of each area. We can also use the function `sm_centroid` to compute
the lon and lat coordinates of the centroid of each region in our dataset.
This is useful for quickly plotting large spatial datasets without needing
all the machinery of the spatial geometries (which are great, but can be
slow at scale).

```{r}
dept %>%
  mutate(sm_centroid(geometry))
```

We can also use the `st_transform` function to *project* the coordinates in our
dataset into a coordinate system to best plot our data. Each coordinate system
uses a numeric code called is [EPSG](https://epsg.io/) code; you can look up the
best one for the region you are interested in. For example, for Metropolitain
France we might use [EPSG:3943](https://epsg.io/3943). This can be done with the
following code.

```{r}
dept %>%
  slice(1:96) %>%
  st_transform(3943) %>%
  ggplot() +
    geom_sf()
```

Note that this create a slightly more accurate map of the data and that the
y-axis and lines of longitude are no-longer parallel.

Likewise, for Guadeloupe, Martinique and Guyane we might use
[EPSG:2972](https://epsg.io/2972):

```{r}
dept %>%
  slice(97:99) %>%
  st_transform(2972) %>%
  ggplot() +
    geom_sf(alpha = 0.1) +
    geom_sf_text(aes(label = departement_name), size = 2)
```

And for La Réunion and Mayotte we might use [EPSG:5879](https://epsg.io/5879):

```{r}
dept %>%
  slice(100:101) %>%
  st_transform(5879) %>%
  ggplot() +
    geom_sf(alpha = 0.1) +
    geom_sf_text(aes(label = departement_name), size = 2)
```

We will see that these projections help particularly when using large regions
like the U.S.; they are also particularly useful for projecting data near the
North or South Pole.

## Practice

### Map projections

To get a better sense of how map projections work, slice the `dept` data to
include only rows 96-101. This gives one département in Europe and all of the
DOM. Plot the data using both `geom_sf` and `geom_sf_text` (for a label, use
`departement_name`). Do not project the data, so that it uses just longitude
and latitude. Note that Val-d'Oise is in Europe.

```{r, question-01, warning = FALSE}

```

Take the code from above and use the EPSG-3943 projection; this was
recommend for use with European France. Notice that the European
part of the map has the lat- and lon- lines the closest to parallel to
the x- and y-axes.

```{r, question-02}

```

Finally, use the EPSG-5879 projection with the same data:

```{r, question-03}

```

Note how distorted the rest of the plot becomes, but how nice the lines look
near Mayotte and La Réunion.

### Population Density

Combine the methods in the notes to add a population density variable to the
dataset (people per square kilometer). Plot the data spatially using
color to show the population density for the first 96 rows of the dataset.
Use a viridis color scale and appropropriate projection.

```{r, question-04}

```

You should see that Paris and the the surrounding areas are by far the most
dense areas. This is because they are the only départements that include only a
dense city area and not any of the surrounding countryside. To fix this we need
to be more careful about how we define a color palette.

Usually, we use color only as a secondary element in a plot. However, in a map
we often need to use color to show the main feature of interest. This means that
we have to be fairly careful about how colors are defined. Change the plot
from your previous question to have the following scale:

- `scale_fill_distiller(trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10)`

```{r, question-05}

```

Other options for palettes include:

  - **Diverging:** BrBG, PiYG, PRGn, PuOr, RdBu, RdGy, RdYlBu, RdYlGn, Spectral
  - **Qualitative:** Accent, Dark2, Paired, Pastel1, Pastel2, Set1, Set2, Set3
  - **Sequential:** Blues, BuGn, BuPu, GnBu, Greens, Greys, Oranges, OrRd, PuBu,
                    PuBuGn, PuRd, Purples, RdPu, Reds, YlGn, YlGnBu, YlOrBr, YlOrRd

Keep this in mind as we continue to work with spatial data, as the defaults will
often look terrible and be hard to interpret.

### Creating spatial data

It is time to get a little bit more complicated. Let's read in a dataset of
the largest French cities:

```{r, message=FALSE}
cities <- read_csv(file.path("data", "france_cities.csv"))
cities
```

This is just a normal data frame object, but we can convert it into a
spatial object with the following code:

```{r}
cities <- st_as_sf(cities, coords = c("lon", "lat"), remove = FALSE)
st_crs(cities) <- 4326  # set the default projection; EPSG-4326 => lat/long
cities
```

Unlike the geometries in `dept`, each row here is only a single point rather
than a polygon. We can do spatial scatterplots without dealing with geometry
objects, but need to do this convertion to make projections work correctly.

In the code below, take the last plot you made and add a `geom_sf_label` layer
showing the names and locations of the largest 10 French cities. This helps add
context to the plot for anyone not particularly familiar with the locations of
the major cities (Paris is already easy to see; the others less-so).

```{r, question-06}

```

You can adjust the number of cities based on how large the plot is on your screen.

### Integrating the COVID-19 data

Finally, let's integrate a bit of the COVID-19 data into our spatial analysis.
Start by creating a dataset called `covid_april` by filtering the data to
include only the data from "2020-04-30" and compute the number of COVID-19
deaths per 100k residents. This will require using the `pop` and `covid`
datasets.

```{r, question-07}

```

For the 96 départements in Metropolitain France, draw a map showing the
number of people who have died from COVID-19 per 1000 residents on 30 April.
Consider using the code you had in the previous plot (with a distiller
palette and the top cities shown).

```{r, question-08}

```

As a last step (you will find you can copy much of the previous code),
show the number of people who were hospitalised per 100k residents
on "2020-10-01".

```{r, question-09}

```

If you have time remaining, consider adding some labels to the last two plots to
make them closer to something that you might include a published report. You
might consider adding `theme_void` to minimize the ink used on the axes.
