---
title: "Notebook 11"
output: html_document
---

## Getting Started

Before running this notebook, select "Session > Restart R and Clear Output" in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
```

## Movies Data

Over the next few classes we will be working with a dataset of movies I have
constructed consisting of the top 100 grossing films for each year from 1970 to
2019. The data comes from IMDb. Today we will focus on getting familiar with
the various components of the data. Let's read in the four tables of data,
as well as a data dictionary, and then go through each of the tables.

```{r, message=FALSE}
movies <- read_csv(file.path("data", "movies_50_years.csv"))
m_genre <- read_csv(file.path("data", "movies_50_years_genre.csv"))
m_people <- read_csv(file.path("data", "movies_50_years_people.csv"))
m_dict <- read_csv(file.path("data", "movies_50_years_data_dictionary.csv"))
m_color <- read_csv(file.path("data", "movies_50_years_color.csv"))
```

The movies dataset contains one row for each movie. Most of the variables are
fairly straightforward; you can see the units by opening the data dictionary.
Three variables concern the movie's theatrical poster. These given the average
brightness (average pixel intensity), saturation (are the colors bright or
washed out / white), and complexity (a cartoon image would have a low
complexity; lots of text or icons would have a high complexity).

```{r, message=FALSE}
movies
```

A second dataset gives more detailed information about each poster by indicating
how much of a poster is of a certain color. If you want to look at the movie
poster itself, just search for a film on [IMDb](https://www.imdb.com) and
search for the film. The poster is the first image on the film's page.

```{r, message=FALSE}
m_color
```

We also have a dataset of movie genres. The data structure is straightforward,
but needs to be kept in its own table because a single movies can be assigned
to multiple genres.

```{r, message=FALSE}
m_genre
```

Finally, we also have a dataset of people associated with each film. We do not
have a lot of metadata about the people, but I have added a prediction of each
person's gender based on U.S. Social Security records. These are not always
correct (there is a confidence score included as well) but are useful for some
aggregate analyses.

```{r, message=FALSE}
m_people
```

There is a lot to unpack with these datasets. We will need to make use of the
many methods we have learned so far this semester to make sense of the data.

### Movie Genre

Start by making a bar plot showing the number of times each genre tag appears
in the dataset. No need for fancy labels or titles, but do order the categories
from smallest to largest and consider flipping the axes if your screen is too
narrow to read the vertical-bar version of the plot.

```{r, question-01}

```

Now, we are going to do something a bit more complicated. In the chunk below,
create a visualization that shows which genre tag is used the most in each
year. The plot is easy; creating the dataset will take some work.

```{r, question-02}

```

Finally, we are going to create a dataset that has one row for each film that
can be merged into the main movies dataset. First, create a dataset that
collapses all of the genres for a film into a single value using the function
`sm_paste`.

```{r, question-03}

```

And then, create a dataset that associates each film to the "least popular"
genre associated with it. For example *MASH* is listed as a comedy, drama, and
war film. You should have seen above that the "war" tag is much less common
than "comedy" or "drama", so it should be associated with "war".

```{r, question-04}

```

Assigning the least-popular genre typically gives the most appropriate tag to
each movie because it will usually also be the most specific genre.

### Movie Color

Start by verifying that the percentage values for each film each add up to 100
(there may be some slight rounding error, but everything should be very close
to 100). Use whatever method you find to be the easiest or most reliable, but
do not resort to manually checking the values for each film.

```{r, question-05}

```

Next, associate each film to the color that is most dominent in each film
poster. Plot a count of the most dominant colors using a bar plot. No need for
any labels, titles, or other finishing touches, but do try to color the bars
according to the associated color name. Note that you will have to use the
aesthetic "fill" and scale `scale_fill_identity` to do this.

```{r, question-06}

```
