---
title: "Project 03"
output:
  html_document:
    theme: cosmo
    highlight: zenburn
    css: "note-style.css"
---

## Solutions

This document gives solutions to the questions posed in Project 3. I had a bit
of fun with some of these, and included additional tweaks that make the output
look a bit nicer. These were not expected in my grading of the projects.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)
library(smodels)
library(stringi)
library(lubridate)
library(sf)
library(units)
library(RcppRoll)

theme_set(theme_minimal())
options(dplyr.summarise.inform = FALSE)
options(width = 77L)
options(lubridate.week.start = 1)
Sys.setlocale(locale = "en_US.UTF-8")

sm_centroid <- function(data) {
  suppressWarnings({ z <- st_coordinates(st_centroid(data)) })
  return(tibble(lon = z[,1], lat = z[,2]))
}

county <- read_sf(file.path("data", "us_counties.geojson"))
covid <- read_csv(file.path("data", "us_county_covid.csv"))
demog <- read_csv(file.path("data", "us_county_metadata.csv"))
```

## Prompts

### 1. Spatial Extent

**Question** Show a spatial visualization (i.e., a map) of the infection rate
(cases per population) in the U.S. on 1 May 2020, 15 June 2020, and 1 October
for the "lower 48" states (no Hawaii, Alaska, or Puerto Rico). Also, separately
show the infection rate for Hawaii on 15 June 2020. Try to describe the patterns
in a few short sentences. Make sure to use appropriate projections and color
scales.

**Solution Notes** I included explicit limits in the color scale in order to
make the maps comparable to one another. For Hawaii, I manually set the limits
of the x and y axes to zoom in on the actual data (there must be a better way
to do this automatically, but I could not figure out how). I also included
thick outlines for Hawaii; these are often too busy for counties but look nice
when looking at island chains.

```{r}
county %>%
  inner_join(filter(covid, date == "2020-05-01"), by = c("fips", "state")) %>%
  inner_join(demog, by = c("fips", "state")) %>%
  mutate(infect_rate = cases / population * 100000) %>%
  filter(!(state %in% c("HI", "AK", "PR"))) %>%
  st_transform(3083) %>%
  ggplot() +
    geom_sf(aes(fill = infect_rate), size = 0) +
    scale_fill_distiller(
      trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10,
      limits = c(2, 8192)
    ) +
    labs(fill = "Infection Rate\n (per 100k)") +
    theme_void()
```

```{r}
county %>%
  inner_join(filter(covid, date == "2020-06-15"), by = c("fips", "state")) %>%
  inner_join(demog, by = c("fips", "state")) %>%
  mutate(infect_rate = cases / population * 100000) %>%
  filter(!(state %in% c("HI", "AK", "PR"))) %>%
  st_transform(3083) %>%
  ggplot() +
    geom_sf(aes(fill = infect_rate), size = 0) +
    scale_fill_distiller(
      trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10,
      limits = c(2, 8192)
    ) +
    labs(fill = "Infection Rate\n (per 100k)") +
    theme_void()
```

```{r}
county %>%
  inner_join(filter(covid, date == "2020-10-01"), by = c("fips", "state")) %>%
  inner_join(demog, by = c("fips", "state")) %>%
  mutate(infect_rate = cases / population * 100000) %>%
  filter(!(state %in% c("HI", "AK", "PR"))) %>%
  st_transform(3083) %>%
  ggplot() +
    geom_sf(aes(fill = infect_rate), size = 0) +
    scale_fill_distiller(
      trans = "log2", palette = "Spectral", guide = "legend", n.breaks = 10,
      limits = c(2, 8192)
    ) +
    labs(fill = "Infection Rate\n (per 100k)") +
    theme_void()
```

```{r}
county %>%
  inner_join(filter(covid, date == "2020-06-15"), by = c("fips", "state")) %>%
  inner_join(demog, by = c("fips", "state")) %>%
  mutate(infect_rate = cases / population * 100000) %>%
  filter(state == "HI") %>%
  st_transform(26961) %>%
  ggplot() +
    geom_sf(aes(fill = infect_rate), size = 0.7, color = "black") +
    scale_fill_distiller(
      palette = "Spectral", guide = "legend", n.breaks = 10
    ) +
    labs(fill = "Infection Rate\n (per 100k)") +
    coord_sf(xlim = c(0, 572762.8), ylim = c(8500, 400000)) +
    theme_void()

```

### 2. Political Party

**Question** How does the general political affiliation at the county level
relate to the observed case fatality rate (cases / deaths) and the overall
infection rate (cases / population)? Answer the question on a month-by-month
basis (i.e., compute the rates for each month separately, not cumulatively) and
plot these showing the change over time.

**Solution Notes** Several students asked a good question about whether we
should compute the infection and fatality rates for each county and then take
an average, or add together the deaths and cases and compute the rates on the
entire data. The later is the better choice because it correctly weights the
counties by their overall population. A small technical detail is that I
filtered the data to only the first day of the month and then used the lead
function, so that I have the counts for the ongoing month. I did it this way
because R puts the month label on the first day of the month. Depending on your
approach, you plot may be off-set by a month in relation to the labels on the
x-axis.

```{r}
party_data <- covid %>%
  filter(day(date) == 1) %>%
  inner_join(demog, by = c("fips", "state")) %>%
  group_by(fips) %>%
  mutate(
    deaths_new = lead(deaths) - deaths,
    cases_new = lead(cases) - cases,
    party = if_else(per_gop_2016 > per_dem_2016, "R", "D")
  ) %>%
  filter(!is.na(cases_new), !is.na(deaths_new)) %>%
  group_by(date, party) %>%
  summarize(
    n = n(),
    inf_rate = sum(cases_new) / sum(population),
    case_fat_rate = sum(deaths_new) / sum(cases_new)
  )
```


```{r}
party_data %>%
  ggplot(aes(date, case_fat_rate)) +
    geom_line(aes(color = party)) +
    scale_color_manual(values = c("blue", "red"))
```

```{r}
party_data %>%
  ggplot(aes(date, inf_rate)) +
    geom_line(aes(color = party)) +
    scale_color_manual(values = c("blue", "red"))
```


### 3. Largest Counties

**Question** Visualize the number of new cases in the 8 largest counties over
the time range of the entire dataset for each day using a 7-day moving average.

**Solution Notes** I had to change the original question because the 9th largest
county is Kings County (Brooklyn). New York only reports city-wide data, so
this caused missing values to appear in the data. In my solution, I added on a
layer we never learned called a facet wrap. This create a number of small
plots, broken out by a new variable. Do not overuse this feature (that is why
I never showed it in class), but it can produce very nice visualizations in
the right situations.

```{r}
top_counties <- demog %>%
  arrange(desc(population)) %>%
  slice(1:8, 10)   # I will add in the 10th largest, to make the plot look nice

covid %>%
  semi_join(top_counties, by = c("fips", "state")) %>%
  group_by(fips) %>%
  mutate(cases_new = cases - lag(cases)) %>%
  mutate(cases_new_smooth = roll_meanr(cases_new, n = 7)) %>%
  filter(!is.na(cases_new_smooth)) %>%
  ggplot(aes(date, cases_new_smooth)) +
    geom_line(aes(color = county), show.legend = FALSE) +
    scale_x_date(date_breaks = "month", date_labels = "%b") +
    theme_sm() +
    facet_wrap(~county)
```

### 4. Population Density

**Question** Compute the population density of each county. Group the counties
into 5 groups based on their population density and compute the overall fatality
rate (cases / deaths), infection rate (cases / population) and mortality rate
(deaths / cases) for each group on 1 October.

**Solution Notes** The `cut` function here is key to making the solution
compact. I had originally expected you to come up with manual breaks, but that
was probably too aggressive, in retrospect. The `cut` function takes the range
of the data (minimum to maximum) and splits it into equally sized buckets. That
is, the range of each bucket is equally size, not the amount of data in each.
The density distribution has a heavy tail, so we need to take a log transform
to make better choices about the cutoffs.

```{r}
county_density <- county %>%
  left_join(demog, by = c("fips", "state")) %>%
  mutate(area = as.numeric(set_units(st_area(geometry), "km^2"))) %>%
  mutate(density = population / area) %>%
  mutate(bucket = cut(log2(density + 1), breaks = 5, labels = FALSE)) %>%
  as_tibble()

covid %>%
  filter(date == "2020-10-01") %>%
  inner_join(county_density, by = c("fips", "state")) %>%
  group_by(bucket) %>%
  summarize(
    case_fatality_rate = sum(deaths) / sum(cases),
    infection_rate = sum(cases) / sum(population),
    total_mortality_rate = sum(deaths) / sum(population)
  )
```


### 5. Cases and Deaths Lag

**Question** At the county level, aggregated over each week, what's the "lag"
between deaths and cases? That is, if we look at the number of deaths in one
week in the dataset is it most closely correlated with the number of cases in
the week in question, one week before, two weeks before, three weeks before,
or more? As an optional challenge, try to plot the relationship without
resorting to manually copying the numbers (you might need a pivot function
here).

**Solution Notes** We would expect that the number of deaths in a county lead
the number of cases. That is, the spike in cases usually comes several weeks
before the spike in deaths. The solution here follows what I was expecting you
to do for the question. Some students included a default value for the lags;
I would not suggest this (how do you know there were zero cases outside the
range of the data? I am fairly sure this is quite false, particularly for the
counties hardest hit in the early months of the pandemic), but I did not take
off any points for this.

```{r}
covid %>%
  filter(wday(date) == 1) %>%
  group_by(fips, state) %>%
  mutate(cases_new = cases - lag(cases), deaths_new = deaths - lag(deaths)) %>%
  mutate(
    cases_new_0 = cases_new,
    cases_new_1 = lag(cases_new, n = 1),
    cases_new_2 = lag(cases_new, n = 2),
    cases_new_3 = lag(cases_new, n = 3),
    cases_new_4 = lag(cases_new, n = 4),
    cases_new_5 = lag(cases_new, n = 5),
    cases_new_6 = lag(cases_new, n = 6),
    cases_new_7 = lag(cases_new, n = 7)
  ) %>%
  ungroup() %>%
  summarise(
    sm_cor(deaths_new, cases_new_0),
    sm_cor(deaths_new, cases_new_1),
    sm_cor(deaths_new, cases_new_2),
    sm_cor(deaths_new, cases_new_3),
    sm_cor(deaths_new, cases_new_4),
    sm_cor(deaths_new, cases_new_5),
    sm_cor(deaths_new, cases_new_6),
    sm_cor(deaths_new, cases_new_7)
  ) %>%
  pivot_longer(cols = everything()) %>%
  mutate(week_lag = as.numeric(stri_extract(name, regex = "[0-9]"))) %>%
  ggplot(aes(week_lag, value)) +
    geom_line()
```

**Bonus** The code below is an modification/extension of question 5 that
includes three changes: it uses 7-day rolling averages, computes the daily lags
rather than weekly lags, and uses a weighted correlation as a function of the
population. To do this without copying a lot of code, I use a for loop to cycle
over each of the lags. In both plots you can see that deaths lag behind cases
by about 3-4 weeks.

```{r}
covid_new <- covid %>%
  group_by(fips, state) %>%
  mutate(cases_new = cases - lag(cases), deaths_new = deaths - lag(deaths)) %>%
  mutate(
    cases_new = roll_meanr(cases_new, n = 7),
    deaths_new = roll_meanr(deaths_new, n = 7),
  ) %>%
  inner_join(demog, by = c("fips", "state")) %>%
  group_by(fips)

lag_data <- tibble(day_lag = seq(0, 60), cor = NA_real_)
for (j in seq_len(nrow(lag_data)))
{
  covid_new <- mutate(covid_new, cases_new_lag = lag(cases_new, n = j - 1))

  X <- as.matrix(select(ungroup(covid_new), deaths_new, cases_new_lag, population))
  X <- X[apply(!is.na(X), 1, all),]
  lag_data$cor[j] <- cov.wt(X[, 1:2], wt = X[, 3], cor = TRUE)$cor[1, 2]
}

lag_data %>%
  ggplot(aes(day_lag, cor)) +
    geom_point() +
    geom_line() +
    scale_x_continuous(breaks = seq(0, 60, by = 5), minor_breaks = NULL)
```
